---
title: "project_clean"
output: html_notebook
---


```{r}
# Load required packges:
suppressPackageStartupMessages({
library(dplyr)
library(Seurat)
})
```

```{r}
# setwd("~/Users/adriana/Documents/Stanford_Classes/First Year/winter 2026/stemrem 205/stemrem")
setwd("/Users/ryannayebi/Downloads/STANFORD/win25/BMDS205/205_proj")
```

### rahcel & perla cleaned up this code so that it started with what i want 

```{r}
metadata.DVC <- read.table("data/metadata_all_DVC.txt", sep = "\t", header = TRUE)
metadata.DVC <- metadata.DVC[-1,]
head(metadata.DVC)

allmice.data <- readRDS("data/allmice.rds")
head(allmice.data)
```

## GSEA Setup

### Packages and pipeline settings
```{r gsea-packages-and-settings}
project_r_libs <- normalizePath("r_libs", mustWork = FALSE)
.libPaths(c(project_r_libs, .libPaths()))

suppressPackageStartupMessages({
  library(jsonlite)
  library(digest)
  library(fgsea)
  library(tibble)
  library(ggplot2)
})

dir.create("cache", showWarnings = FALSE, recursive = TRUE)
Sys.setenv(R_USER_CACHE_DIR = normalizePath("cache", mustWork = FALSE))

gsea_pipeline_settings <- list(
  output_root = "results/gsea",
  summary_dir = "results/gsea/_summary",
  ranking_method = "avg_log2FC",
  de_method_label = "Seurat FindMarkers (per-cell, exploratory)",
  de_params = list(
    test.use = "wilcox",
    slot = "data",
    fc.slot = "data",
    pseudocount.use = 1,
    logfc.threshold = 0.1,
    min.pct = 0.01,
    min.diff.pct = -Inf,
    max.cells.per.ident = Inf,
    only.pos = FALSE,
    latent.vars = NULL,
    min.cells.feature = 3,
    min.cells.group = 3,
    densify = FALSE
  ),
  fgsea_params = list(
    minSize = 10,
    maxSize = 500,
    seed = 123
  ),
  curve_thresholds = list(
    padj_max = 0.25,
    abs_nes_min = 1.5
  ),
  msigdb = list(
    db_species = "MM",
    collection = "MH",
    species = "Mus musculus"
  )
)
```

### Contrast registry
```{r gsea-contrast-registry}
build_contrast_registry <- function() {
  tibble::tibble(
    contrast_id = c(
      "all_glp1r_vs_gfral",
      "neurons_glp1r_vs_gfral",
      "all_adlib_vs_fasted",
      "all_refed_vs_fasted",
      "all_adlib_vs_refed",
      "all_neuron_vs_glial",
      "neurons_adlib_vs_fasted",
      "neurons_refed_vs_fasted",
      "glia_glp1r_vs_gfral",
      "glia_adlib_vs_fasted",
      "glia_refed_vs_fasted"
    ),
    contrast_type = c(
      "receptor_within_population",
      "receptor_within_population",
      "diet_within_population",
      "diet_within_population",
      "diet_within_population",
      "class_vs_class",
      "diet_within_population",
      "diet_within_population",
      "receptor_within_population",
      "diet_within_population",
      "diet_within_population"
    ),
    object_name = c(
      "filtered_Glp1r_Gfral",
      "neurons",
      "filtered_Glp1r_Gfral",
      "filtered_Glp1r_Gfral",
      "filtered_Glp1r_Gfral",
      "filtered_Glp1r_Gfral",
      "neurons",
      "neurons",
      "glia",
      "glia",
      "glia"
    ),
    group_by = c(
      "cell_type",
      "cell_type",
      "diet_category",
      "diet_category",
      "diet_category",
      "identity_layer1",
      "diet_category",
      "diet_category",
      "cell_type",
      "diet_category",
      "diet_category"
    ),
    ident1 = c(
      "Glp1r",
      "Glp1r",
      "Ad libitum",
      "Refed",
      "Ad libitum",
      "neuron",
      "Ad libitum",
      "Refed",
      "Glp1r",
      "Ad libitum",
      "Refed"
    ),
    ident2 = c(
      "Gfral",
      "Gfral",
      "Fasted",
      "Fasted",
      "Refed",
      "glial",
      "Fasted",
      "Fasted",
      "Gfral",
      "Fasted",
      "Fasted"
    ),
    notes = c(
      "Receptor comparison in all receptor-positive cells",
      "Receptor comparison in neurons only",
      "Diet comparison in all receptor-positive cells",
      "Diet comparison in all receptor-positive cells",
      "Diet comparison in all receptor-positive cells",
      "Class comparison (neuron vs glial)",
      "Diet comparison in neurons",
      "Diet comparison in neurons",
      "Receptor comparison in glia",
      "Diet comparison in glia",
      "Diet comparison in glia"
    )
  )
}
```

### Object preparation helpers
```{r gsea-object-helpers}
derive_cell_type <- function(obj) {
  receptor_expr <- FetchData(obj, vars = c("Glp1r", "Gfral"))
  obj$cell_type <- dplyr::case_when(
    receptor_expr$Glp1r > 0 & receptor_expr$Gfral > 0 ~ "Both",
    receptor_expr$Glp1r > 0 ~ "Glp1r",
    receptor_expr$Gfral > 0 ~ "Gfral",
    TRUE ~ "Neither"
  )
  obj
}

derive_diet_category <- function(obj) {
  obj$diet_category <- dplyr::case_when(
    obj$orig.ident == "adlib" ~ "Ad libitum",
    obj$orig.ident == "fast" ~ "Fasted",
    obj$orig.ident == "refed" ~ "Refed",
    TRUE ~ as.character(obj$orig.ident)
  )
  obj
}

prepare_gsea_object <- function(obj) {
  obj <- derive_cell_type(obj)
  derive_diet_category(obj)
}

build_gsea_analysis_objects <- function(filtered_Glp1r_Gfral, neurons) {
  base_objects <- list(
    filtered_Glp1r_Gfral = filtered_Glp1r_Gfral,
    neurons = neurons,
    glia = subset(filtered_Glp1r_Gfral, subset = identity_layer1 == "glial")
  )

  lapply(base_objects, prepare_gsea_object)
}

resolve_analysis_object <- function(analysis_objects, name) {
  analysis_objects[[name]]
}
```

### Cache and DEG helpers
```{r gsea-cache-and-deg-helpers}
read_json_if_present <- function(path) {
  if (file.exists(path)) {
    jsonlite::fromJSON(path, simplifyVector = FALSE)
  } else {
    list(key_hash = NULL)
  }
}

write_json_pretty <- function(x, path) {
  jsonlite::write_json(x, path = path, pretty = TRUE, auto_unbox = TRUE, null = "null")
}

file_md5_or_na <- function(path) {
  if (file.exists(path)) {
    unname(tools::md5sum(path))
  } else {
    NA_character_
  }
}

validate_group_idents <- function(obj, group_by, ident1, ident2) {
  group_counts <- table(as.character(obj@meta.data[[group_by]]))
  list(
    n_cells_total = length(Cells(obj)),
    n_cells_ident1 = as.integer(group_counts[ident1]),
    n_cells_ident2 = as.integer(group_counts[ident2]),
    group_counts = as.list(group_counts)
  )
}

build_de_key <- function(obj, contrast_row, de_params, validation_info, de_method_label) {
  list(
    contrast_id = contrast_row$contrast_id,
    object_name = contrast_row$object_name,
    group_by = contrast_row$group_by,
    ident1 = contrast_row$ident1,
    ident2 = contrast_row$ident2,
    seurat_version = as.character(packageVersion("Seurat")),
    assay_used = DefaultAssay(obj),
    de_params = de_params,
    n_cells_total = validation_info$n_cells_total,
    n_cells_ident1 = validation_info$n_cells_ident1,
    n_cells_ident2 = validation_info$n_cells_ident2,
    cell_barcode_hash = digest::digest(sort(Cells(obj)), algo = "xxhash64"),
    group_counts_hash = digest::digest(validation_info$group_counts, algo = "xxhash64"),
    de_method_label = de_method_label
  )
}

compute_or_load_deg <- function(obj, contrast_row, contrast_dir, settings) {
  dir.create(contrast_dir, showWarnings = FALSE, recursive = TRUE)
  deg_rds <- file.path(contrast_dir, "deg_results.rds")
  deg_csv <- file.path(contrast_dir, "deg_results.csv")
  deg_meta_json <- file.path(contrast_dir, "deg_meta.json")

  validation_info <- validate_group_idents(
    obj = obj,
    group_by = contrast_row$group_by,
    ident1 = contrast_row$ident1,
    ident2 = contrast_row$ident2
  )

  de_key <- build_de_key(
    obj = obj,
    contrast_row = contrast_row,
    de_params = settings$de_params,
    validation_info = validation_info,
    de_method_label = settings$de_method_label
  )
  de_key_hash <- digest::digest(de_key, algo = "xxhash64")

  prior_meta <- read_json_if_present(deg_meta_json)
  cache_hit <- file.exists(deg_rds) && identical(prior_meta$key_hash, de_key_hash)
  cache_status <- if (cache_hit) "hit" else if (file.exists(deg_rds)) "mismatch_recompute" else "miss_recompute"

  if (cache_hit) {
    deg_df <- readRDS(deg_rds)
  } else {
    deg_df <- FindMarkers(
      object = obj,
      ident.1 = contrast_row$ident1,
      ident.2 = contrast_row$ident2,
      group.by = contrast_row$group_by,
      test.use = settings$de_params$test.use,
      slot = settings$de_params$slot,
      fc.slot = settings$de_params$fc.slot,
      pseudocount.use = settings$de_params$pseudocount.use,
      logfc.threshold = settings$de_params$logfc.threshold,
      min.pct = settings$de_params$min.pct,
      min.diff.pct = settings$de_params$min.diff.pct,
      max.cells.per.ident = settings$de_params$max.cells.per.ident,
      only.pos = settings$de_params$only.pos,
      latent.vars = settings$de_params$latent.vars,
      min.cells.feature = settings$de_params$min.cells.feature,
      min.cells.group = settings$de_params$min.cells.group,
      densify = settings$de_params$densify
    )
    saveRDS(deg_df, deg_rds)

    deg_csv_tbl <- tibble::rownames_to_column(as.data.frame(deg_df), var = "gene")
    write.csv(deg_csv_tbl, deg_csv, row.names = FALSE)

    write_json_pretty(
      list(
        generated_at = as.character(Sys.time()),
        key_hash = de_key_hash,
        key = de_key,
        cache_status = cache_status
      ),
      deg_meta_json
    )
  }

  list(
    deg_df = deg_df,
    cache_status = cache_status,
    key_hash = de_key_hash,
    validation = validation_info,
    files = list(
      deg_rds = deg_rds,
      deg_csv = deg_csv,
      deg_meta_json = deg_meta_json
    )
  )
}
```

### Ranking and fgsea helpers
```{r gsea-ranking-and-fgsea-helpers}
build_ranks <- function(deg_df) {
  ranked_raw <- tibble::rownames_to_column(as.data.frame(deg_df), var = "gene") %>%
    dplyr::filter(!is.na(gene), gene != "", is.finite(avg_log2FC)) %>%
    dplyr::mutate(rank_score = avg_log2FC)

  ranked_tbl <- ranked_raw %>%
    dplyr::arrange(dplyr::desc(abs(rank_score))) %>%
    dplyr::distinct(gene, .keep_all = TRUE)

  ranks <- ranked_tbl$rank_score
  names(ranks) <- ranked_tbl$gene
  ranks <- sort(ranks, decreasing = TRUE)

  list(
    ranked_tbl = ranked_tbl,
    ranks = ranks,
    n_ranked_genes = nrow(ranked_raw),
    n_unique_ranked_genes = nrow(ranked_tbl)
  )
}

find_msigdb_cache_file <- function(cache_root) {
  cache_dir <- file.path(cache_root, "R", "msigdbr")
  if (!dir.exists(cache_dir)) {
    return(NA_character_)
  }

  candidates <- list.files(cache_dir, pattern = "msigdb.*\\.rds$", full.names = TRUE)
  if (length(candidates) == 0) {
    return(NA_character_)
  }

  candidates[which.max(file.info(candidates)$mtime)]
}

run_or_load_fgsea <- function(ranks, hallmark_pathways, contrast_row, contrast_dir, settings, msigdb_info) {
  fgsea_rds <- file.path(contrast_dir, "hallmark_fgsea.rds")
  fgsea_csv <- file.path(contrast_dir, "hallmark_fgsea.csv")
  fgsea_meta_json <- file.path(contrast_dir, "fgsea_meta.json")

  fgsea_key <- list(
    contrast_id = contrast_row$contrast_id,
    ranking_method = settings$ranking_method,
    msigdb_db_species = settings$msigdb$db_species,
    msigdb_collection = settings$msigdb$collection,
    msigdb_cache_file = msigdb_info$cache_file,
    msigdb_cache_md5 = msigdb_info$cache_md5,
    fgsea_params = settings$fgsea_params,
    rank_vector_hash = digest::digest(ranks, algo="xxhash64")
  )
  fgsea_key_hash <- digest::digest(fgsea_key, algo="xxhash64")

  prior_meta <- read_json_if_present(fgsea_meta_json)
  cache_hit <- file.exists(fgsea_rds) && identical(prior_meta$key_hash, fgsea_key_hash)
  cache_status <- if (cache_hit) "hit" else if (file.exists(fgsea_rds)) "mismatch_recompute" else "miss_recompute"

  if (cache_hit) {
    gsea_df <- readRDS(fgsea_rds)
  } else {
    set.seed(settings$fgsea_params$seed)
    gsea_df <- fgsea::fgseaMultilevel(
      pathways = hallmark_pathways,
      stats = ranks,
      minSize = settings$fgsea_params$minSize,
      maxSize = settings$fgsea_params$maxSize
    )
    gsea_df <- as.data.frame(gsea_df) %>%
      dplyr::arrange(padj, dplyr::desc(abs(NES)))

    saveRDS(gsea_df, fgsea_rds)

    gsea_csv_tbl <- gsea_df %>%
      dplyr::mutate(
        leadingEdge = vapply(leadingEdge, paste, collapse = ";", FUN.VALUE = character(1))
      )
    write.csv(gsea_csv_tbl, fgsea_csv, row.names=FALSE)

    write_json_pretty(
      list(
        generated_at = as.character(Sys.time()),
        key_hash = fgsea_key_hash,
        key = fgsea_key,
        cache_status = cache_status
      ),
      fgsea_meta_json
    )
  }

  list(
    gsea_df = gsea_df,
    cache_status = cache_status,
    key_hash = fgsea_key_hash,
    files = list(
      fgsea_rds = fgsea_rds,
      fgsea_csv = fgsea_csv,
      fgsea_meta_json = fgsea_meta_json
    )
  )
}

compute_match_metrics <- function(ranks, hallmark_pathways, gsea_df) {
  pathway_genes <- unique(unlist(hallmark_pathways))
  n_ranked <- length(ranks)
  n_matched <- length(intersect(names(ranks), pathway_genes))
  match_rate <- n_matched/max(n_ranked, 1)

  list(
    n_pathways_library = length(hallmark_pathways),
    n_pathways_tested = nrow(gsea_df),
    n_genes_matched_to_pathways = n_matched,
    match_rate = match_rate,
    low_match_warning = match_rate < 0.5
  )
}
```

### Plotting and manifest helpers
```{r gsea-plotting-and-manifest-helpers}
make_summary_plots <- function(gsea_df, contrast_row, contrast_dir) {
  if (nrow(gsea_df) == 0) {
    return(list(bar_plot = NA_character_, bubble_plot = NA_character_))
  }

  plot_tbl <- gsea_df %>%
    dplyr::filter(is.finite(NES), is.finite(padj)) %>%
    dplyr::arrange(padj, dplyr::desc(abs(NES)), pathway) %>%
    dplyr::slice_head(n = 10)

  if (nrow(plot_tbl) == 0) {
    return(list(bar_plot = NA_character_, bubble_plot = NA_character_))
  }

  plot_tbl <- plot_tbl %>%
    dplyr::mutate(
      pathway = factor(pathway, levels = pathway[order(NES)]),
      direction = ifelse(NES > 0, paste0("Higher in ", contrast_row$ident1), paste0("Higher in ", contrast_row$ident2)),
      neg_log10_padj = -log10(pmax(padj, 1e-300))
    )

  contrast_label <- paste0(contrast_row$contrast_id, ": ", contrast_row$ident1, " vs ", contrast_row$ident2)
  bar_path <- file.path(contrast_dir, "hallmark_top_nes_bar.png")
  bubble_path <- file.path(contrast_dir, "hallmark_bubble.png")

  gsea_bar <- ggplot(plot_tbl, aes(x = pathway, y = NES, fill = NES > 0)) +
    geom_col() +
    coord_flip() +
    scale_fill_manual(values = c("#3C78D8", "#CC4125"), guide = "none") +
    labs(
      title = paste("Hallmark GSEA:", contrast_label),
      x = NULL,
      y = "Normalized Enrichment Score (NES)"
    ) +
    theme_minimal()
  ggsave(filename = bar_path, plot = gsea_bar, width = 8, height = 6, dpi = 300)

  gsea_bubble <- ggplot(plot_tbl, aes(x = NES, y = pathway, size = neg_log10_padj, color = direction)) +
    geom_point(alpha = 0.85) +
    scale_color_manual(values = stats::setNames(
      object = c("#3C78D8", "#CC4125"),
      nm = c(paste0("Higher in ", contrast_row$ident2), paste0("Higher in ", contrast_row$ident1))
    )) +
    labs(
      title = paste0("Hallmark GSEA Summary:\n", contrast_label),
      subtitle = "Point size = -log10(adjusted p-value)",
      x = "Normalized Enrichment Score (NES)",
      y = NULL,
      color = "Direction",
      size = "-log10(padj)"
    ) +
    theme_minimal()
  ggsave(filename = bubble_path, plot = gsea_bubble, width = 8.5, height = 6, dpi = 300)

  list(bar_plot = bar_path, bubble_plot = bubble_path)
}

make_curve_plots_if_signal <- function(gsea_df, ranks, hallmark_pathways, contrast_row, contrast_dir, thresholds) {
  if (nrow(gsea_df) == 0) {
    return(list(
      positive_curve = NA_character_,
      negative_curve = NA_character_,
      curves_generated = FALSE,
      curve_plots_skipped_reason = "No pathways tested",
      min_padj = Inf,
      top_abs_nes = 0
    ))
  }

  min_padj <- suppressWarnings(min(gsea_df$padj, na.rm = TRUE))
  if (!is.finite(min_padj)) min_padj <- Inf

  top_abs_nes <- suppressWarnings(max(abs(gsea_df$NES), na.rm = TRUE))
  if (!is.finite(top_abs_nes)) top_abs_nes <- 0

  gate_pass <- (min_padj < thresholds$padj_max) || (top_abs_nes > thresholds$abs_nes_min)
  if (!gate_pass) {
    reason <- paste0(
      "Gated off: min(padj)=", signif(min_padj, 3),
      " and top|NES|=", signif(top_abs_nes, 3),
      " did not pass thresholds."
    )
    return(list(
      positive_curve = NA_character_,
      negative_curve = NA_character_,
      curves_generated = FALSE,
      curve_plots_skipped_reason = reason,
      min_padj = min_padj,
      top_abs_nes = top_abs_nes
    ))
  }

  positive_path <- NA_character_
  negative_path <- NA_character_

  top_positive <- gsea_df %>%
    dplyr::filter(NES > 0) %>%
    dplyr::slice_min(order_by = padj, n = 1)

  top_negative <- gsea_df %>%
    dplyr::filter(NES < 0) %>%
    dplyr::slice_min(order_by = padj, n = 1)

  if (nrow(top_positive) > 0) {
    positive_name <- top_positive$pathway[[1]]
    p_positive <- fgsea::plotEnrichment(hallmark_pathways[[positive_name]], ranks) + ggtitle(
      paste0("Top Positive Trend: ", positive_name),
      paste0("NES=", round(top_positive$NES[[1]], 2), ", padj=", signif(top_positive$padj[[1]], 3))
    )
    positive_path <- file.path(contrast_dir, "top_positive_curve.png")
    ggsave(filename = positive_path, plot = p_positive, width = 8, height = 5, dpi = 300)
  }

  if (nrow(top_negative) > 0) {
    negative_name <- top_negative$pathway[[1]]
    p_negative <- fgsea::plotEnrichment(hallmark_pathways[[negative_name]], ranks) + ggtitle(
      paste0("Top Negative Trend: ", negative_name),
      paste0("NES=", round(top_negative$NES[[1]], 2), ", padj=", signif(top_negative$padj[[1]], 3))
    )
    negative_path <- file.path(contrast_dir, "top_negative_curve.png")
    ggsave(filename = negative_path, plot = p_negative, width = 8, height = 5, dpi = 300)
  }

  list(
    positive_curve = positive_path,
    negative_curve = negative_path,
    curves_generated = !is.na(positive_path) || !is.na(negative_path),
    curve_plots_skipped_reason = NA_character_,
    min_padj = min_padj,
    top_abs_nes = top_abs_nes
  )
}

load_hallmark_pathways <- function(settings) {
  cache_file <- find_msigdb_cache_file(Sys.getenv("R_USER_CACHE_DIR"))
  if (is.na(cache_file)) {
    stop("MSigDB cache file not found in cache/R/msigdbr.")
  }

  hallmark_tbl <- readRDS(cache_file) %>%
    dplyr::filter(
      gs_collection == settings$msigdb$collection,
      db_target_species == settings$msigdb$db_species
    ) %>%
    dplyr::transmute(
      gs_name = gs_name,
      gene_symbol = db_gene_symbol
    ) %>%
    dplyr::filter(!is.na(gene_symbol), gene_symbol != "") %>%
    dplyr::distinct()

  pathways <- split(hallmark_tbl$gene_symbol, hallmark_tbl$gs_name)

  list(
    pathways = pathways,
    cache_file = cache_file,
    cache_md5 = file_md5_or_na(cache_file)
  )
}

write_contrast_manifest_row <- function(
  contrast_row, started_at, finished_at,
  deg_result, rank_result, fgsea_result, match_metrics, curve_result,
  settings = gsea_pipeline_settings
) {
  tibble::tibble(
    contrast_id = contrast_row$contrast_id,
    contrast_type = contrast_row$contrast_type,
    object_name = contrast_row$object_name,
    group_by = contrast_row$group_by,
    ident1 = contrast_row$ident1,
    ident2 = contrast_row$ident2,
    status = "success",
    error_message = NA_character_,
    started_at = as.character(started_at),
    finished_at = as.character(finished_at),
    de_cache_status = deg_result$cache_status,
    fgsea_cache_status = fgsea_result$cache_status,
    de_key_hash = deg_result$key_hash,
    fgsea_key_hash = fgsea_result$key_hash,
    n_cells_total = deg_result$validation$n_cells_total,
    n_cells_ident1 = deg_result$validation$n_cells_ident1,
    n_cells_ident2 = deg_result$validation$n_cells_ident2,
    n_ranked_genes = rank_result$n_ranked_genes,
    n_unique_ranked_genes = rank_result$n_unique_ranked_genes,
    n_pathways_library = match_metrics$n_pathways_library,
    n_pathways_tested = match_metrics$n_pathways_tested,
    n_genes_matched_to_pathways = match_metrics$n_genes_matched_to_pathways,
    match_rate = match_metrics$match_rate,
    low_match_warning = match_metrics$low_match_warning,
    min_padj = curve_result$min_padj,
    top_abs_nes = curve_result$top_abs_nes,
    curves_generated = curve_result$curves_generated,
    curve_plots_skipped_reason = curve_result$curve_plots_skipped_reason,
    ranking_method = settings$ranking_method,
    de_method_label = settings$de_method_label
  )
}

collect_top_pathways <- function(gsea_df, contrast_row) {
  gsea_df %>%
    dplyr::arrange(padj, dplyr::desc(abs(NES))) %>%
    dplyr::slice_head(n = 10) %>%
    dplyr::mutate(
      contrast_id = contrast_row$contrast_id,
      contrast_type = contrast_row$contrast_type,
      ident1 = contrast_row$ident1,
      ident2 = contrast_row$ident2,
      rank_within_contrast = dplyr::row_number(),
      direction = ifelse(NES > 0, contrast_row$ident1, contrast_row$ident2)
    ) %>%
    dplyr::select(
      contrast_id, contrast_type, ident1, ident2, rank_within_contrast,
      pathway, NES, padj, size, direction
    )
}
```

##START HERE FOR PROJECT

##rachel helped looking for low quality: mt 
## the chunk below works to reload after deleted allmice.data from environemnt 
```{r}
# library(Seurat)
# allmice.data <- readRDS("/Users/adriana/Documents/Stanford Classes/First Year/winter 2026/stemrem 205/stemrem1/stemrem/allmice.rds")
# head(allmice.data)
# metadata <- read.csv('metadata.DVC.csv')
# head(metadata)
```

```{r}
allmice.data[["percent.mt"]] <- PercentageFeatureSet(allmice.data, pattern = "^MT-")
table(allmice.data$percent.mt)
```


```{r}
allmice.data[["percent.mt"]] <- PercentageFeatureSet(allmice.data, pattern = "^(mt-|MT-)")
table(allmice.data$percent.mt)
```

#very few genes to mt 
```{r}
VlnPlot(allmice.data, features = c( "percent.mt"), ncol = 1)
```


```{r}
features <- rownames(allmice.data)
head(features)
```


#looking to make sure these exist 
```{r}
gene_names <- c("Glp1r", "Gfral")
present_genes <- gene_names[gene_names %in% rownames(allmice.data)]
print(present_genes)
```

```{r}
gene_expression <- FetchData(allmice.data, vars = c("Glp1r", "Gfral"))
head(gene_expression)
```

##i made an rds file with all the mouse stuff and sorted for glp1r and gfral. but i did not sort mitochondria or features so may need to redo!!! (its we checked mitchondria and it seems like the authors already sorted that )
##this output is GLP1 and GRFAL positive CELLS (not neurons, so any cells with at least one glp1r or gfral)
#defining threshold as greater than 0
```{r}
threshold <- 0
# Identify cells where at least one of the genes exceeds the threshold
cells_to_keep <- rownames(gene_expression)[gene_expression$Glp1r > threshold | gene_expression$Gfral > threshold]
filtered_Glp1r_Gfral <- allmice.data[, cells_to_keep]
length(Cells(filtered_Glp1r_Gfral)) 
saveRDS(filtered_Glp1r_Gfral, "filtered_Glp1r_Gfral.rds")
```

##filter to only Glp1r and not gfral, this output is only glp1 
```{r}
threshold <- 0
# Identify cells where at least one of the genes exceeds the threshold (| = or)
Glp1r_cells_to_keep <- rownames(gene_expression)[gene_expression$Glp1r > threshold & gene_expression$Gfral == 0]
filtered_Glp1r <- allmice.data[, Glp1r_cells_to_keep]
length(Cells(filtered_Glp1r)) 
saveRDS(filtered_Glp1r, "filtered_Glp1r.rds")
```

##filter to only gfral, this output is only gfral
```{r}
threshold <- 0
# Identify cells where at least one of the genes exceeds the threshold (Gfral only)
Gfral_cells_to_keep <- rownames(gene_expression)[gene_expression$Glp1r == 0 & gene_expression$Gfral > threshold]
filtered_Gfral <- allmice.data[, Gfral_cells_to_keep]
length(Cells(filtered_Gfral)) 
saveRDS(filtered_Gfral, "filtered_Gfral.rds")
```

```{r}
library(dplyr)
```

#normal set with both and each separate
```{r}
# normalize
filtered_Glp1r_Gfral <- NormalizeData(filtered_Glp1r_Gfral)
filtered_Gfral <- NormalizeData(filtered_Gfral)
filtered_Glp1r <- NormalizeData(filtered_Glp1r)
```

#look for variable genes in combined with both glp1 and gfral 
```{r}
# identify variable genes in the group (this is not separating them)
#genes with the most variance in the dataset 
filtered_Glp1r_Gfral <- FindVariableFeatures(filtered_Glp1r_Gfral)
```
```{r}
receptor_expression <- FetchData(filtered_Glp1r_Gfral, vars = c("Glp1r", "Gfral"))
head(receptor_expression)
```

#variable genes in separate
```{r}
##identify variable genes between these 
filtered_Glp1r <- FindVariableFeatures(filtered_Glp1r)
filtered_Gfral <- FindVariableFeatures(filtered_Gfral)
```

# top 10 for each separated 
```{r}
# view top 10
top10_Glp1r <- head(VariableFeatures(filtered_Glp1r), 10)
top10_Gfral <- head(VariableFeatures(filtered_Gfral), 10)
top10_Glp1r
top10_Gfral
```

#plot each separately

```{r}
# plot variablility distribution
plot1 <- VariableFeaturePlot(filtered_Glp1r)
plot_Glp1r <- LabelPoints(plot = plot1, points = top10_Glp1r, repel = TRUE)
plot2 <- VariableFeaturePlot(filtered_Gfral)
plot_Gfral <- LabelPoints(plot = plot2, points = top10_Gfral, repel = TRUE) 
plot_Glp1r
plot_Gfral
```


```{r}
# Z-score scale (genes)
Glp1r.Gfral.genes <- rownames(filtered_Glp1r_Gfral)
Glp1r.genes <- rownames(filtered_Glp1r)
filtered_Glp1r <- ScaleData(filtered_Glp1r, features = Glp1r.genes)
Gfral.genes <- rownames(filtered_Gfral)
filtered_Gfral <- ScaleData(filtered_Gfral, features = Gfral.genes)
filtered_Glp1r_Gfral <- ScaleData(filtered_Glp1r_Gfral, features = Glp1r.Gfral.genes)
```

```{r}
head(filtered_Glp1r@assays$originalexp$counts)
head(filtered_Glp1r@meta.data)
head(filtered_Gfral@assays$originalexp$counts)
head(filtered_Gfral@meta.data)
```

# PCA, default subsets to highly variable features
#output shows positive and negative genes
```{r}
filtered_Glp1r <- RunPCA(filtered_Glp1r)
filtered_Gfral <- RunPCA(filtered_Gfral)
filtered_Glp1r_Gfral <- RunPCA(filtered_Glp1r_Gfral)
```

###how do i spread out the cells more so they arent all on top of each other like this? 
fixed I think right @adriana?
```{r}
# A visualization that is useful for determining how many PCs to include
# pick a dot for number of PCAs
ElbowPlot(filtered_Glp1r)
ElbowPlot(filtered_Gfral)
ElbowPlot(filtered_Glp1r_Gfral)
```

# computing the neighborhood graph and find clusters for each receptor type and also all my cells 
#higher resolution is more clusters
##starting with 10 
```{r}
filtered_Glp1r <- FindNeighbors(filtered_Glp1r, dims = 1:5)
filtered_Glp1r <- FindClusters(filtered_Glp1r, resolution = 0.15)
filtered_Gfral <- FindNeighbors(filtered_Gfral, dims = 1:4)
filtered_Gfral <- FindClusters(filtered_Gfral, resolution = 0.15)
filtered_Glp1r_Gfral <- FindNeighbors(filtered_Glp1r_Gfral, dims = 1:6)
filtered_Glp1r_Gfral <- FindClusters(filtered_Glp1r_Gfral, resolution = 0.15)
```

#UMAP of Glp1r alone, Gfral alone, and combined subset
```{r}
filtered_Glp1r <- RunUMAP(filtered_Glp1r, dims = 1:6)
filtered_Gfral <- RunUMAP(filtered_Gfral, dims = 1:6)
filtered_Glp1r_Gfral <- RunUMAP(filtered_Glp1r_Gfral, dims = 1:6)
```

# plot UMAP
```{r}
DimPlot(filtered_Glp1r, reduction = "umap")
DimPlot(filtered_Gfral, reduction = "umap")
DimPlot(filtered_Glp1r_Gfral, reduction = "umap")
```


## trying to annotate clusters
```{r}
cluster_membership <- filtered_Glp1r_Gfral$seurat_clusters
table(cluster_membership)
```

### we use findallmarkers and find DIFFERENCES between clusters
##FindAllMarkersidentifies markers for all clusters in a Seurat object relative to all other clusters, differential expression analysis across all clusters 
##output is a data frame that includes a row for each gene, annotated by the cluster where it is significantly expressed
```{r}
cluster_markers <- FindAllMarkers(filtered_Glp1r_Gfral)
# View top markers for each cluster
# head(cluster_markers)
```

bookmark
##significant cluster markers
```{r}
significant_markers <- cluster_markers %>%
  filter(p_val_adj < 0.05 & abs(avg_log2FC) > 1) 
head(significant_markers)
dim(significant_markers)
```
```{r}
library(dplyr)

# Assuming your data is named 'markers' and has 'gene' and 'cluster' columns
genes_by_cluster <- significant_markers %>%
  group_by(cluster) %>%
  summarize(genes = list(gene))  # Create a list of genes for each cluster

# View the result
print(genes_by_cluster)
```

### look at top different genes by cluster
```{r}

#cluster0
sorted_cluster0_genes <- significant_markers %>%
  filter(cluster == "0") %>%
  arrange(desc(avg_log2FC)) %>%
  select(gene, avg_log2FC)


# View the genes for cluster0
print(sorted_cluster0_genes)

#cluster 1
sorted_cluster1_genes <- significant_markers %>%
  filter(cluster == "1") %>%
  arrange(desc(avg_log2FC)) %>%
  select(gene, avg_log2FC)

print(sorted_cluster1_genes)

#cluster 2
sorted_cluster2_genes <- significant_markers %>%
  filter(cluster == "2") %>%
  arrange(desc(avg_log2FC)) %>%
  select(gene, avg_log2FC)


print(sorted_cluster2_genes)

#cluster3
sorted_cluster3_genes <- significant_markers %>%
  filter(cluster == "3") %>%
  arrange(desc(avg_log2FC)) %>%
  select(gene, avg_log2FC)

print(sorted_cluster3_genes)

#cluster4
sorted_cluster4_genes <- significant_markers %>%
  filter(cluster == "4") %>%
  arrange(desc(avg_log2FC)) %>%
  select(gene, avg_log2FC)

print(sorted_cluster4_genes)

#cluster5
sorted_cluster5_genes <- significant_markers %>%
  filter(cluster == "5") %>%
  arrange(desc(avg_log2FC)) %>%
  select(gene, avg_log2FC)

print(sorted_cluster5_genes)

```

###I want to look at top genes for every cluster, trying to identify identity overlap !!!
##trying to figure out top genes for each cluster, not just the different ones
##after running the code again now it only has 8 clusters 0-7. finding clusters again probably caused the issue! (need to use elbow plot to guide, if these clusters all look the same w top cells, try lower elbow plot #)
# 0.15 res has 6 clusters seeing if this helps
#0.17 had 7 clusters

```{r}
library(Seurat)
library(dplyr)

# Ensure clustering has been performed on your data
#filtered_Glp1r_Gfral <- FindClusters(filtered_Glp1r_Gfral)

# Get the average expression per cluster
average_expression <- AverageExpression(filtered_Glp1r_Gfral, return.seurat = FALSE)

# Convert the average expression to a data frame
avg_expr_df <- as.data.frame(average_expression$originalexp)

# Create an empty list to hold top genes for each cluster
top_genes_list <- list()

# Specify the number of top genes to retrieve
top_n <- 10

# Loop through each column (cluster) to find top genes
for (cluster in colnames(avg_expr_df)) {
  # Get the top N genes for the cluster
  top_genes <- avg_expr_df %>%
    arrange(desc(!!sym(cluster))) %>%  # Sort the genes by their average expression
    head(top_n) %>%                    # Select the top N genes
    rownames()                         # Extract gene names
  top_genes_list[[cluster]] <- top_genes  # Store results in the list
}

# Combine the results into a data frame for easier viewing
top_genes_df <- do.call(rbind, lapply(names(top_genes_list), function(cluster_name) {
  data.frame(Cluster = cluster_name, Gene = top_genes_list[[cluster_name]], stringsAsFactors = FALSE)
}))

# View the results
print(top_genes_df)
```

## separating each column is one cluster !!!
## this gives us gene is row and expr for each cluster
```{r}
library(Seurat)
library(dplyr)
library(tidyr)
library(tibble)  # For rownames_to_column()

# Ensure clustering has been performed on your data
#filtered_Glp1r_Gfral <- FindClusters(filtered_Glp1r_Gfral) ## DO NOT REPEAT THIS KEEP MY DIMS

# Get the average expression per cluster
average_expression <- AverageExpression(filtered_Glp1r_Gfral, return.seurat = FALSE)

# Convert the average expression to a data frame
avg_expr_df <- as.data.frame(average_expression$originalexp)

# Check the dim and structure of the average expression data frame
print(dim(avg_expr_df))  # Expect genes x clusters
print(head(avg_expr_df))  # Preview the first few rows

# Create an empty list to hold top genes for each cluster
top_genes_list <- list()

# Specify the number of top genes to retrieve
top_n <- 10

# Loop through each column (cluster) to find top genes along with their expression levels
for (cluster in colnames(avg_expr_df)) {
  # Get the top N genes for the cluster
  top_genes <- avg_expr_df %>%
    arrange(desc(!!sym(cluster))) %>%     # Sort the genes by their average expression
    head(top_n) %>%                       # Select the top N genes
    rownames_to_column(var = "Gene") %>%  # Convert row names to a column called "Gene"
    mutate(Cluster = cluster,              # Add a column for the cluster
           Expression = !!sym(cluster))    # Capture the corresponding expression value
  
  # Store results in the list
  top_genes_list[[cluster]] <- top_genes  
}

# Combine the results into a single data frame
top_genes_df <- bind_rows(top_genes_list)

# Pivot the data frame to have genes as rows and clusters as columns
top_genes_wide <- top_genes_df %>%
  select(Gene, Cluster, Expression) %>%  # Ensure the relevant columns are selected
  pivot_wider(names_from = Cluster, values_from = Expression, values_fill = 0)  # Convert to wide format

# View the reshaped data frame
print(top_genes_wide)
```

##this orders by a specified cluster 
```{r}
#ordered by 0
ordered_genes_cluster0 <- top_genes_wide %>%
  arrange(desc(g0))  
print(ordered_genes_cluster0)

#ordered by cluster 1
ordered_genes_cluster1 <- top_genes_wide %>%
  arrange(desc(g1))  
print(ordered_genes_cluster1)

#ordered by cluster 2
ordered_genes_cluster2 <- top_genes_wide %>%
  arrange(desc(g2))  
print(ordered_genes_cluster2)

#ordered by cluster 3
ordered_genes_cluster3 <- top_genes_wide %>%
  arrange(desc(g3))  
print(ordered_genes_cluster3)

#ordered by cluster 4
ordered_genes_cluster4 <- top_genes_wide %>%
  arrange(desc(g4))  
print(ordered_genes_cluster4)

#ordered by cluster 5
ordered_genes_cluster5 <- top_genes_wide %>%
  arrange(desc(g5))  
print(ordered_genes_cluster5)

```


##trying to see which "identity_layer1" most common in cluster

```{r}
data <- filtered_Glp1r_Gfral@meta.data %>%
  select(seurat_clusters, identity_layer1)  

# Find the most frequent identity per cluster
result <- data %>%
  group_by(seurat_clusters) %>%
  summarise(most_frequent_identity1 = names(which.max(table(identity_layer1))))

# View the result
print(result)
```


##trying identity_layer2 - NO REPEATS WITH 0.15 res, so we like 0.15 res better than 0.2 , 6 total clusters and label by identity layer 2, but astrocytes get messed up. trying other res . 0.17 has repeats so stick with 0.15!
```{r}
data2 <- filtered_Glp1r_Gfral@meta.data %>%
  select(seurat_clusters, identity_layer2)  # Make sure 'identity' is the correct name for the identity column

# Find the most frequent identity per cluster
result2 <- data2 %>%
  group_by(seurat_clusters) %>%
  summarise(most_frequent_identity2 = names(which.max(table(identity_layer2))))

# View the result
print(result2)

#how much of each cluster is id2
result2 <- data2 %>%
  group_by(seurat_clusters, identity_layer2) %>%
  summarise(count = n()) %>%
  ungroup() %>%
  group_by(seurat_clusters) %>%
  mutate(proportion = count / sum(count)) %>%
  select(seurat_clusters, identity_layer2, proportion) %>%
  arrange(seurat_clusters, desc(proportion))

# View the result
print(result2)
```


##trying identity_layer3
```{r}
data3 <- filtered_Glp1r_Gfral@meta.data %>%
  select(seurat_clusters, identity_layer3)  # Make sure 'identity' is the correct name for the identity column

# Find the most frequent identity per cluster
result3 <- data3 %>%
  group_by(seurat_clusters) %>%
  summarise(most_frequent_identity3 = names(which.max(table(identity_layer3))))

##how much of each cluster has this identity3
result3 <- data3 %>%
  group_by(seurat_clusters, identity_layer3) %>%
  summarise(count = n()) %>%
  ungroup() %>%
  group_by(seurat_clusters) %>%
  mutate(proportion = count / sum(count)) %>%
  select(seurat_clusters, identity_layer3, proportion) %>%
  arrange(seurat_clusters, desc(proportion))

# View the result
print(result3)
```

## i like it better with 0.15 res ! keeping this
##rename clusters by identity_layer3

```{r}
# 0 = mixed_neurons (because it has a lot of 1. 2 and 3) (layer 3 but more specific than layer 2 i think)
# 1 = Sall3_class (layer 2, more accurate, 70%)
# 2 = M0 (M0 good, 98%)
# 3 = Vglut1 (good, 85%)
# 4 = astrocyte (this is mixed but lets keep it?? figure out bio relevance- golli neurons express glia genes!!!)
# 5 = oligodendrocyte (98%)
# 6 = 

new.cluster.ids <- c("mixed_neurons", 'Sall3_class', "M0", 
                     'Vglut1', 'astrocyte',
                     'oligodendrocyte'
                     )
names(new.cluster.ids) <- levels(filtered_Glp1r_Gfral)
filtered_Glp1r_Gfral <- RenameIdents(filtered_Glp1r_Gfral, new.cluster.ids)
```

```{r}
# plot umap with renamed clusters
DimPlot(filtered_Glp1r_Gfral, reduction = "umap", label = TRUE, pt.size = 0.3)
```

### Looking for cluster identities for filtered_glp1r
##trying to see which "identity_layer1" most common in cluster

```{r}
glp1r_data <- filtered_Glp1r@meta.data %>%
  select(seurat_clusters, identity_layer1)  

# Find the most frequent identity per cluster
glp1r_result <- glp1r_data %>%
  group_by(seurat_clusters) %>%
  summarise(glp1r_most_frequent_identity1 = names(which.max(table(identity_layer1))))

# View the result
print(glp1r_result)
```


##trying identity_layer2 - NO REPEATS WITH 0.15 res, so we like 0.15 res better than 0.2 , 6 total clusters and label by identity layer 2, but astrocytes get messed up. trying other res . 0.17 has repeats so stick with 0.15!
```{r}
glp1r_data2 <- filtered_Glp1r@meta.data %>%
  select(seurat_clusters, identity_layer2)  # Make sure 'identity' is the correct name for the identity column

# Find the most frequent identity per cluster
glp1r_result2 <- glp1r_data2 %>%
  group_by(seurat_clusters) %>%
  summarise(glp1r_most_frequent_identity2 = names(which.max(table(identity_layer2))))

# View the result
print(glp1r_result2)

#how much of each cluster is id2
glp1r_result2 <- glp1r_data2 %>%
  group_by(seurat_clusters, identity_layer2) %>%
  summarise(count = n()) %>%
  ungroup() %>%
  group_by(seurat_clusters) %>%
  mutate(proportion = count / sum(count)) %>%
  select(seurat_clusters, identity_layer2, proportion) %>%
  arrange(seurat_clusters, desc(proportion))

# View the result
print(glp1r_result2)
```


##trying identity_layer3
```{r}
glp1r_data3 <- filtered_Glp1r@meta.data %>%
  select(seurat_clusters, identity_layer3)  # Make sure 'identity' is the correct name for the identity column

# Find the most frequent identity per cluster
glp1r_result3 <- glp1r_data3 %>%
  group_by(seurat_clusters) %>%
  summarise(glp1r_most_frequent_identity3 = names(which.max(table(identity_layer3))))
print(glp1r_result3)

##how much of each cluster has this identity3
glp1r_result3 <- glp1r_data3 %>%
  group_by(seurat_clusters, identity_layer3) %>%
  summarise(count = n()) %>%
  ungroup() %>%
  group_by(seurat_clusters) %>%
  mutate(proportion = count / sum(count)) %>%
  select(seurat_clusters, identity_layer3, proportion) %>%
  arrange(seurat_clusters, desc(proportion))

# View the result
print(glp1r_result3)
```

##looking at interesting genes on all of the umaps
```{r}
genes_to_plot = c( 'Glp1r', 'Gfral',  'Vglut' )
FeaturePlot(filtered_Glp1r, reduction = "umap", features = genes_to_plot)
FeaturePlot(filtered_Gfral, reduction = "umap", features = genes_to_plot)
FeaturePlot(filtered_Glp1r_Gfral, reduction = "umap", features = genes_to_plot)
```

##NOW MOVING ON TO DEGS

## compare DEGs between glp1r and gfral cells using the combined dataset 

#identifying cells to pick
```{r}
glp1r_cells <- WhichCells(filtered_Glp1r_Gfral, expression = Glp1r > 0)  # Adjust the threshold based on your needs
gfral_cells <- WhichCells(filtered_Glp1r_Gfral, expression = Gfral > 0) 
```

#this makes a column called cell_type and labels cells as both, glp1r, or gfral so we can compare only glp1r and gfral
##FindMarkers compares our two groups and the output is DEGs between the two groups
```{r}
filtered_Glp1r_Gfral <- AddMetaData(filtered_Glp1r_Gfral, 
                                     metadata = ifelse(Cells(filtered_Glp1r_Gfral) %in% glp1r_cells & 
                                                       Cells(filtered_Glp1r_Gfral) %in% gfral_cells, 
                                                       "Both", 
                                                       ifelse(Cells(filtered_Glp1r_Gfral) %in% glp1r_cells, 
                                                              "Glp1r", 
                                                              "Gfral")), 
                                     col.name = "cell_type")

# Identify differentially expressed genes between the two cell types of interest, this selected the cells we want and not BOTH
deg_results <- FindMarkers(filtered_Glp1r_Gfral, ident.1 = "Glp1r", ident.2 = "Gfral", group.by = "cell_type")

# View the results
## this provides q value by BH
head(deg_results)
```

#sort the degs 
```{r}
deg_results_sorted <- deg_results[order(deg_results$avg_log2FC), ]
head(deg_results_sorted)
significant_genes <- deg_results[deg_results$p_val_adj < 0.05 & abs(deg_results$avg_log2FC) > 1, ]
head(significant_genes)
dim(deg_results_sorted)
dim(significant_genes)
print(significant_genes)
```

```{r}
# Create a new data frame for plotting
library(ggplot2)

# Add a column for -log10(p_val_adj) for plotting
deg_results$log2FoldChange <- deg_results$avg_log2FC
deg_results$negLog10PValue <- -log10(deg_results$p_val_adj)

# Remove NA values if any (they can cause plotting issues)
deg_results <- na.omit(deg_results)
```

##volcano with enhanced volcano
```{r}
BiocManager::install("EnhancedVolcano", force = TRUE)
library(EnhancedVolcano)
```

##this is showing us DEGs, first plot shows glp1r and gfral then we narrow it down to see the other genes 
```{r}
# Create the volcano plot
EnhancedVolcano(deg_results,
    lab = rownames(deg_results),
    x = 'log2FoldChange',
    y = 'p_val_adj',                        
    pCutoff = 0.05, )

EnhancedVolcano(deg_results,
    lab = rownames(deg_results),
    x = 'log2FoldChange',
    y = 'p_val_adj',                        
    pCutoff = 0.05,
    xlim = c(-5, 5),                          
    ylim = c(0, 10), )
```

### Now we are subsetting to ONLY neurons. repeat the above DEG analysis from the whole cell population with only neurons with our receptors. this is neruons only 
#this subsets to only neurons 
```{r}
neurons <- subset(filtered_Glp1r_Gfral, subset = identity_layer1 == "neuron")
head(neurons)
saveRDS(neurons, "filtered_Glp1r_Gfral_neurons.rds")
```

##looking at degs between glp1r and gfral neurons only 

```{r}
# Identify differentially expressed genes between the two cell types of interest 
neuron_deg_results <- FindMarkers(neurons, ident.1 = "Glp1r", ident.2 = "Gfral", group.by = "cell_type")

# View the results
## this provides q value by BH
head(neuron_deg_results)
```

```{r}
neuron_deg_results_sorted <- neuron_deg_results[order(neuron_deg_results$avg_log2FC), ]
head(neuron_deg_results_sorted)
significant_genes_neuron <- neuron_deg_results[neuron_deg_results$p_val_adj < 0.05 & abs(neuron_deg_results$avg_log2FC) > 1, ]
head(significant_genes_neuron)
dim(neuron_deg_results_sorted)
dim(significant_genes_neuron)
print(significant_genes_neuron)
```
```{r}
neuron_deg_results$log2FoldChange <- neuron_deg_results$avg_log2FC
neuron_deg_results$negLog10PValue <- -log10(neuron_deg_results$p_val_adj)

# Remove NA values if any (they can cause plotting issues)
neuron_deg_results <- na.omit(neuron_deg_results)
```

```{r}
# Create the volcano plot
EnhancedVolcano(neuron_deg_results,
    lab = rownames(neuron_deg_results),
    x = 'log2FoldChange',
    y = 'p_val_adj',                        
    pCutoff = 0.05, )

EnhancedVolcano(neuron_deg_results,
    lab = rownames(neuron_deg_results),
    x = 'log2FoldChange',
    y = 'p_val_adj',                        
    pCutoff = 0.05,
    xlim = c(-5, 5),                          
    ylim = c(0, 10), )
```

## Registry Driven GSEA Pipeline
All pairwise FindMarkers contrasts
### Registry inputs
```{r gsea-registry-inputs}
row_to_list <- function(row_tbl) {
  list(
    contrast_id = row_tbl$contrast_id[[1]],
    contrast_type = row_tbl$contrast_type[[1]],
    object_name = row_tbl$object_name[[1]],
    group_by = row_tbl$group_by[[1]],
    ident1 = row_tbl$ident1[[1]],
    ident2 = row_tbl$ident2[[1]],
    notes = row_tbl$notes[[1]]
  )
}

dir.create(gsea_pipeline_settings$output_root, showWarnings = FALSE, recursive = TRUE)
dir.create(gsea_pipeline_settings$summary_dir, showWarnings = FALSE, recursive = TRUE)

gsea_analysis_objects <- build_gsea_analysis_objects(
  filtered_Glp1r_Gfral = filtered_Glp1r_Gfral,
  neurons = neurons
)

hallmark_info <- load_hallmark_pathways(gsea_pipeline_settings)
contrast_registry <- build_contrast_registry()
```

### One-contrast runner
```{r gsea-single-contrast-runner}
run_single_gsea_contrast <- function(contrast_row, analysis_objects, hallmark_info, settings) {
  contrast_dir <- file.path(settings$output_root, contrast_row$contrast_id)
  dir.create(contrast_dir, showWarnings = FALSE, recursive = TRUE)

  started_at <- Sys.time()
  obj <- resolve_analysis_object(analysis_objects, contrast_row$object_name)

  deg_result <- compute_or_load_deg(
    obj = obj,
    contrast_row = contrast_row,
    contrast_dir = contrast_dir,
    settings = settings
  )

  rank_result <- build_ranks(deg_result$deg_df)
  write.csv(
    rank_result$ranked_tbl,
    file.path(contrast_dir, "ranked_genes.csv"),
    row.names = FALSE
  )

  fgsea_result <- run_or_load_fgsea(
    ranks = rank_result$ranks,
    hallmark_pathways = hallmark_info$pathways,
    contrast_row = contrast_row,
    contrast_dir = contrast_dir,
    settings = settings,
    msigdb_info = hallmark_info
  )

  match_metrics <- compute_match_metrics(
    ranks = rank_result$ranks,
    hallmark_pathways = hallmark_info$pathways,
    gsea_df = fgsea_result$gsea_df
  )

  if (isTRUE(match_metrics$low_match_warning)) {
    warning(
      paste0(
        "Low pathway match rate for ", contrast_row$contrast_id,
        ": ", signif(match_metrics$match_rate, 3)
      )
    )
  }

  make_summary_plots(
    gsea_df = fgsea_result$gsea_df,
    contrast_row = contrast_row,
    contrast_dir = contrast_dir
  )

  curve_result <- make_curve_plots_if_signal(
    gsea_df = fgsea_result$gsea_df,
    ranks = rank_result$ranks,
    hallmark_pathways = hallmark_info$pathways,
    contrast_row = contrast_row,
    contrast_dir = contrast_dir,
    thresholds = settings$curve_thresholds
  )

  finished_at <- Sys.time()

  list(
    manifest_row = write_contrast_manifest_row(
      contrast_row = contrast_row,
      started_at = started_at,
      finished_at = finished_at,
      deg_result = deg_result,
      rank_result = rank_result,
      fgsea_result = fgsea_result,
      match_metrics = match_metrics,
      curve_result = curve_result,
      settings = settings
    ),
    top_pathways = collect_top_pathways(fgsea_result$gsea_df, contrast_row)
  )
}
```

### Run all contrasts
```{r gsea-run-all-contrasts}
contrast_results <- lapply(seq_len(nrow(contrast_registry)), function(i) {
  contrast_row <- row_to_list(contrast_registry[i, ])
  run_single_gsea_contrast(
    contrast_row = contrast_row,
    analysis_objects = gsea_analysis_objects,
    hallmark_info = hallmark_info,
    settings = gsea_pipeline_settings
  )
})

manifest_rows <- lapply(contrast_results, function(result) result$manifest_row)
top_pathway_rows <- lapply(contrast_results, function(result) result$top_pathways)
```

### Write summary tables
```{r gsea-write-summary-tables}
run_manifest <- dplyr::bind_rows(manifest_rows) %>%
  dplyr::arrange(contrast_type, contrast_id)
write.csv(
  run_manifest,
  file.path(gsea_pipeline_settings$summary_dir, "run_manifest.csv"),
  row.names = FALSE
)

top_pathways_by_contrast <- dplyr::bind_rows(top_pathway_rows) %>%
  dplyr::arrange(contrast_type, contrast_id, rank_within_contrast)
write.csv(
  top_pathways_by_contrast,
  file.path(gsea_pipeline_settings$summary_dir, "top_pathways_by_contrast.csv"),
  row.names = FALSE
)

top_pathways_by_group <- top_pathways_by_contrast %>%
  dplyr::group_by(contrast_type, pathway) %>%
  dplyr::summarise(
    n_contrasts = dplyr::n_distinct(contrast_id),
    contrasts = paste(sort(unique(contrast_id)), collapse = ";"),
    best_padj = min(padj, na.rm = TRUE),
    median_padj = median(padj, na.rm = TRUE),
    mean_NES = mean(NES, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  dplyr::arrange(contrast_type, best_padj, dplyr::desc(n_contrasts), dplyr::desc(abs(mean_NES)))
write.csv(
  top_pathways_by_group,
  file.path(gsea_pipeline_settings$summary_dir, "top_pathways_by_group.csv"),
  row.names = FALSE
)
```

### Manifest preview
```{r gsea-manifest-preview}
run_manifest %>%
  dplyr::select(
    contrast_id, status, de_cache_status, fgsea_cache_status,
    n_ranked_genes, n_pathways_tested, match_rate,
    curves_generated, curve_plots_skipped_reason
  )
```


### now looking between adlib vs fasted (can do this with any feeding condition) from our all cells. we may want to do this with neurons and glia separately 

###DEGS for ad libitum vs fast in our subset !!! 
## i think this pattern can be used for like neurons and stuff too. NEED to read paper to see what is already done! but i think this way of subsetting is unique 
```{r}
distinct_values <- unique(filtered_Glp1r_Gfral$orig.ident)
print(distinct_values)
```


#creating a new column with the name diet_category like we did with cell_type (THIS IS NOT SPECIFIC TO NEURONS!)
```{r}
filtered_Glp1r_Gfral <- AddMetaData(
  object = filtered_Glp1r_Gfral,
  metadata = case_when(
    filtered_Glp1r_Gfral$orig.ident == "adlib" ~ "Ad libitum",
    filtered_Glp1r_Gfral$orig.ident == "fast" ~ "Fasted",
    filtered_Glp1r_Gfral$orig.ident == "refed" ~ "Refed",  
    TRUE ~ "Other"  # Fallback for any other conditions not specifically captured
  ),
  col.name = "diet_category"
)

head(filtered_Glp1r_Gfral)
# Create a frequency table for the 'diet' column
diet_counts <- table(filtered_Glp1r_Gfral$diet_category)

# Print the counts
print(diet_counts)
```

```{r}
# Identify differentially expressed genes between the diet types
diet_deg_results <- FindMarkers(filtered_Glp1r_Gfral, ident.1 = "Ad libitum", ident.2 = "Fasted", group.by = "diet_category")

# View the results
## this provides q value by BH
head(diet_deg_results)
```

```{r}
head(diet_deg_results)
```

```{r}
diet_deg_results_sorted <- diet_deg_results[order(diet_deg_results$avg_log2FC), ]
head(diet_deg_results_sorted)
significant_genes_diet <- diet_deg_results[diet_deg_results$p_val_adj < 0.05 & abs(diet_deg_results$avg_log2FC) > 1, ]
head(significant_genes_diet)
dim(diet_deg_results_sorted)
dim(significant_genes_diet)
print(significant_genes_diet)
```
```{r}
diet_deg_results$log2FoldChange <- diet_deg_results$avg_log2FC
diet_deg_results$negLog10PValue <- -log10(diet_deg_results$p_val_adj)

# Remove NA values if any (they can cause plotting issues)
diet_deg_results <- na.omit(diet_deg_results)
```

```{r}
# Create the volcano plot
EnhancedVolcano(diet_deg_results,
    lab = rownames(diet_deg_results),
    x = 'log2FoldChange',
    y = 'p_val_adj',                        
    pCutoff = 0.05, )

EnhancedVolcano(diet_deg_results,
    lab = rownames(diet_deg_results),
    x = 'log2FoldChange',
    y = 'p_val_adj',                        
    pCutoff = 0.05,
    xlim = c(-5, 5),                          
    ylim = c(0, 10), )
```

### diet refed vs fast in all glp1r&gfral

```{r}
# Identify differentially expressed genes between the diet types
diet_deg_results2 <- FindMarkers(filtered_Glp1r_Gfral, ident.1 = "Refed", ident.2 = "Fasted", group.by = "diet_category")

# View the results
## this provides q value by BH
head(diet_deg_results2)
```

```{r}
diet_deg_results_sorted2 <- diet_deg_results2[order(diet_deg_results2$avg_log2FC), ]
head(diet_deg_results_sorted2)
significant_genes_diet2 <- diet_deg_results2[diet_deg_results2$p_val_adj < 0.05 & abs(diet_deg_results2$avg_log2FC) > 1, ]
head(significant_genes_diet2)
dim(diet_deg_results_sorted2)
dim(significant_genes_diet2)
print(significant_genes_diet2)
```
```{r}
diet_deg_results2$log2FoldChange <- diet_deg_results2$avg_log2FC
diet_deg_results2$negLog10PValue <- -log10(diet_deg_results2$p_val_adj)

# Remove NA values if any (they can cause plotting issues)
diet_deg_results2 <- na.omit(diet_deg_results2)
```

```{r}
# Create the volcano plot
EnhancedVolcano(diet_deg_results2,
    lab = rownames(diet_deg_results2),
    x = 'log2FoldChange',
    y = 'p_val_adj',                        
    pCutoff = 0.05, )

EnhancedVolcano(diet_deg_results2,
    lab = rownames(diet_deg_results2),
    x = 'log2FoldChange',
    y = 'p_val_adj',                        
    pCutoff = 0.05,
    xlim = c(-5, 5),                          
    ylim = c(0, 10), )
```

## all glp1r and gfral DIET adlib vs refed (bc refed vs fast doesnt seem to have much difference)

```{r}
# Identify differentially expressed genes between the diet types
diet_deg_results3 <- FindMarkers(filtered_Glp1r_Gfral, ident.1 = "Ad libitum", ident.2 = "Refed", group.by = "diet_category")

# View the results
## this provides q value by BH
head(diet_deg_results3)
```

```{r}
diet_deg_results_sorted3 <- diet_deg_results3[order(diet_deg_results3$avg_log2FC), ]
head(diet_deg_results_sorted3)
significant_genes_diet3 <- diet_deg_results3[diet_deg_results3$p_val_adj < 0.05 & abs(diet_deg_results3$avg_log2FC) > 1, ]
head(significant_genes_diet3)
dim(diet_deg_results_sorted3)
dim(significant_genes_diet3)
print(significant_genes_diet3)
```
```{r}
diet_deg_results3$log2FoldChange <- diet_deg_results3$avg_log2FC
diet_deg_results3$negLog10PValue <- -log10(diet_deg_results3$p_val_adj)

# Remove NA values if any (they can cause plotting issues)
diet_deg_results3 <- na.omit(diet_deg_results3)
```

```{r}
# Create the volcano plot
EnhancedVolcano(diet_deg_results3,
    lab = rownames(diet_deg_results3),
    x = 'log2FoldChange',
    y = 'p_val_adj',                        
    pCutoff = 0.05, )

EnhancedVolcano(diet_deg_results3,
    lab = rownames(diet_deg_results3),
    x = 'log2FoldChange',
    y = 'p_val_adj',                        
    pCutoff = 0.05,
    xlim = c(-5, 5),                          
    ylim = c(0, 10), )
```


##trying to label glp1r and gfral in volcano plot for fed vs fasted, they are there but not significant 

```{r}
# Define the genes you want to check
genes_to_check <- c("Glp1r", "Gfral")

# Check if these genes are in the row names of the deg_results
gene_presence <- genes_to_check %in% rownames(diet_deg_results)

# Show which genes are present
print(data.frame(Gene = genes_to_check, Present = gene_presence))
presence_results <- data.frame(Gene = genes_to_check, Present = gene_presence)
print(presence_results)
# Get the rows corresponding to the present genes
present_genes_info <- diet_deg_results[rownames(diet_deg_results) %in% genes_to_check, ]

# Print the information for the present genes
print(present_genes_info)
```

##now looking at DEGs bw neurons and glia in my population
##neurons vs glia identity1 first using both combined 
```{r}
# Perform differential expression analysis comparing neurons and glia
deg_neuron_vs_glia <- FindMarkers(filtered_Glp1r_Gfral, ident.1 = "neuron", ident.2 = "glial", group.by = "identity_layer1")

# View the results
head(deg_neuron_vs_glia)
```
```{r}
deg_neuron_vs_glia$log2FoldChange <- deg_neuron_vs_glia$avg_log2FC
deg_neuron_vs_glia$negLog10PValue <- -log10(deg_neuron_vs_glia$p_val_adj)

# Remove NA values if any (they can cause plotting issues)
deg_neuron_vs_glia <- na.omit(deg_neuron_vs_glia)
```

```{r}
# Create the volcano plot
EnhancedVolcano(deg_neuron_vs_glia,
    lab = rownames(deg_neuron_vs_glia),
    x = 'log2FoldChange',
    y = 'p_val_adj',                        
    pCutoff = 0.05, )

EnhancedVolcano(deg_neuron_vs_glia,
    lab = rownames(deg_neuron_vs_glia),
    x = 'log2FoldChange',
    y = 'p_val_adj',                        
    pCutoff = 0.05,
    xlim = c(-5, 5),                          
    ylim = c(0, 10), )
```


##adlib vs fasted in NEURONS only 

```{r}
# Identify differentially expressed genes between the diet types in NEURONS

neurons <- AddMetaData(
  object = neurons,
  metadata = case_when(
    neurons$orig.ident == "adlib" ~ "Ad libitum",
    neurons$orig.ident == "fast" ~ "Fasted",
    neurons$orig.ident == "refed" ~ "Refed",  
    TRUE ~ "Other"  # Fallback for any other conditions not specifically captured
  ),
  col.name = "diet_category"
)

head(neurons)
# Create a frequency table for the 'diet' column
neurons_diet_counts <- table(neurons$diet_category)

# Print the counts
print(neurons_diet_counts)
```

## neurons adlib vs fast 
```{r}
neurons_diet_deg_results <- FindMarkers(neurons, ident.1 = "Ad libitum", ident.2 = "Fasted", group.by = "diet_category")

# View the results
## this provides q value by BH
head(neurons_diet_deg_results)
```

```{r}
neurons_diet_deg_results_sorted <- neurons_diet_deg_results[order(neurons_diet_deg_results$avg_log2FC), ]
head(neurons_diet_deg_results_sorted)
significant_genes_neurons_diet <- neurons_diet_deg_results[neurons_diet_deg_results$p_val_adj < 0.05 & abs(neurons_diet_deg_results$avg_log2FC) > 1, ]
head(significant_genes_neurons_diet)
dim(neurons_diet_deg_results_sorted)
dim(significant_genes_neurons_diet)
print(significant_genes_neurons_diet)
```
```{r}
neurons_diet_deg_results$log2FoldChange <- neurons_diet_deg_results$avg_log2FC
neurons_diet_deg_results$negLog10PValue <- -log10(neurons_diet_deg_results$p_val_adj)

# Remove NA values if any (they can cause plotting issues)
neurons_diet_deg_results <- na.omit(neurons_diet_deg_results)
```

```{r}
# Create the volcano plot
EnhancedVolcano(neurons_diet_deg_results,
    lab = rownames(neurons_diet_deg_results),
    x = 'log2FoldChange',
    y = 'p_val_adj',                        
    pCutoff = 0.05, )

EnhancedVolcano(neurons_diet_deg_results,
    lab = rownames(neurons_diet_deg_results),
    x = 'log2FoldChange',
    y = 'p_val_adj',                        
    pCutoff = 0.05,
    xlim = c(-5, 5),                          
    ylim = c(0, 10), )
```

### REPEATING WITH FED VS FAST !!!
## neurons refed vs fast this is going to be referred to as diet "2"
```{r}
neurons_diet_deg_results2 <- FindMarkers(neurons, ident.1 = "Refed", ident.2 = "Fasted", group.by = "diet_category")

# View the results
## this provides q value by BH
head(neurons_diet_deg_results2)
```

```{r}
neurons_diet_deg_results_sorted2 <- neurons_diet_deg_results2[order(neurons_diet_deg_results2$avg_log2FC), ]
head(neurons_diet_deg_results_sorted2)
significant_genes_neurons_diet2 <- neurons_diet_deg_results2[neurons_diet_deg_results2$p_val_adj < 0.05 & abs(neurons_diet_deg_results2$avg_log2FC) > 1, ]
head(significant_genes_neurons_diet2)
dim(neurons_diet_deg_results_sorted2)
dim(significant_genes_neurons_diet2)
print(significant_genes_neurons_diet2)
```
```{r}
neurons_diet_deg_results2$log2FoldChange <- neurons_diet_deg_results2$avg_log2FC
neurons_diet_deg_results2$negLog10PValue <- -log10(neurons_diet_deg_results2$p_val_adj)

# Remove NA values if any (they can cause plotting issues)
neurons_diet_deg_results2 <- na.omit(neurons_diet_deg_results2)
```

```{r}
# Create the volcano plot
EnhancedVolcano(neurons_diet_deg_results2,
    lab = rownames(neurons_diet_deg_results2),
    x = 'log2FoldChange',
    y = 'p_val_adj',                        
    pCutoff = 0.05, )

EnhancedVolcano(neurons_diet_deg_results2,
    lab = rownames(neurons_diet_deg_results2),
    x = 'log2FoldChange',
    y = 'p_val_adj',                        
    pCutoff = 0.05,
    xlim = c(-5, 5),                          
    ylim = c(0, 10), )
```
### Now we are subsetting to ONLY GLIA. repeat the above DEG analysis from the whole cell population with only GLIA with our receptors. this is GLIA only 
#SUBSET TO ONLY GLIA 

```{r}
distinct_identity <- unique(filtered_Glp1r_Gfral$identity_layer1)
print(distinct_identity)
```

```{r}
glia <- subset(filtered_Glp1r_Gfral, subset = identity_layer1 == "glial")
head(glia)
saveRDS(glia, "filtered_Glp1r_Gfral_glia.rds")
dim(glia)
```

##GLIA DEGS glp1r and gfral !!!
```{r}
# Identify differentially expressed genes in glia between the two cell types of interest (receptors)
glia_deg_results <- FindMarkers(glia, ident.1 = "Glp1r", ident.2 = "Gfral", group.by = "cell_type")

# View the results
## this provides q value by BH
head(glia_deg_results)
```

```{r}
glia_deg_results_sorted <- glia_deg_results[order(glia_deg_results$avg_log2FC), ]
head(glia_deg_results_sorted)
significant_genes_glia <- glia_deg_results[glia_deg_results$p_val_adj < 0.05 & abs(glia_deg_results$avg_log2FC) > 1, ]
head(significant_genes_glia)
dim(glia_deg_results_sorted)
dim(significant_genes_glia)
print(significant_genes_glia)
```

```{r}
glia_deg_results$log2FoldChange <- glia_deg_results$avg_log2FC
glia_deg_results$negLog10PValue <- -log10(glia_deg_results$p_val_adj)

# Remove NA values if any (they can cause plotting issues)
glia_deg_results <- na.omit(glia_deg_results)
```

```{r}
# Create the volcano plot
EnhancedVolcano(glia_deg_results,
    lab = rownames(glia_deg_results),
    x = 'log2FoldChange',
    y = 'p_val_adj',                        
    pCutoff = 0.05, )

EnhancedVolcano(glia_deg_results,
    lab = rownames(glia_deg_results),
    x = 'log2FoldChange',
    y = 'p_val_adj',                        
    pCutoff = 0.05,
    xlim = c(-5, 5),                          
    ylim = c(0, 10), )
```
### NOW using glia subset do DEGS for adlib vs fasted 
```{r}
# Identify differentially expressed genes between the diet types in GLIA

glia <- AddMetaData(
  object = glia,
  metadata = case_when(
    glia$orig.ident == "adlib" ~ "Ad libitum",
    glia$orig.ident == "fast" ~ "Fasted",
    glia$orig.ident == "refed" ~ "Refed",  
    TRUE ~ "Other"  # Fallback for any other conditions not specifically captured
  ),
  col.name = "diet_category"
)

head(glia)
# Create a frequency table for the 'diet' column
glia_diet_counts <- table(glia$diet_category)

# Print the counts
print(glia_diet_counts)
```

## glia adlib vs fast 
```{r}
glia_diet_deg_results <- FindMarkers(glia, ident.1 = "Ad libitum", ident.2 = "Fasted", group.by = "diet_category")

# View the results
## this provides q value by BH
head(glia_diet_deg_results)
```

```{r}
glia_diet_deg_results_sorted <- glia_diet_deg_results[order(glia_diet_deg_results$avg_log2FC), ]
head(glia_diet_deg_results_sorted)
significant_genes_glia_diet <- glia_diet_deg_results[glia_diet_deg_results$p_val_adj < 0.05 & abs(glia_diet_deg_results$avg_log2FC) > 1, ]
head(significant_genes_glia_diet)
dim(glia_diet_deg_results_sorted)
dim(significant_genes_glia_diet)
print(significant_genes_glia_diet)
```
```{r}
glia_diet_deg_results$log2FoldChange <- glia_diet_deg_results$avg_log2FC
glia_diet_deg_results$negLog10PValue <- -log10(glia_diet_deg_results$p_val_adj)

# Remove NA values if any (they can cause plotting issues)
glia_diet_deg_results <- na.omit(glia_diet_deg_results)
```

```{r}
# Create the volcano plot
EnhancedVolcano(glia_diet_deg_results,
    lab = rownames(glia_diet_deg_results),
    x = 'log2FoldChange',
    y = 'p_val_adj',                        
    pCutoff = 0.05, )

EnhancedVolcano(glia_diet_deg_results,
    lab = rownames(glia_diet_deg_results),
    x = 'log2FoldChange',
    y = 'p_val_adj',                        
    pCutoff = 0.05,
    xlim = c(-5, 5),                          
    ylim = c(0, 10), )
```

####repeat glia degs fed vs fasted 


```{r}
glia_diet_deg_results2 <- FindMarkers(glia, ident.1 = "Refed", ident.2 = "Fasted", group.by = "diet_category")

# View the results
## this provides q value by BH
head(glia_diet_deg_results2)
```

```{r}
glia_diet_deg_results_sorted2 <- glia_diet_deg_results2[order(glia_diet_deg_results2$avg_log2FC), ]
head(glia_diet_deg_results_sorted2)
significant_genes_glia_diet2 <- glia_diet_deg_results2[glia_diet_deg_results2$p_val_adj < 0.05 & abs(glia_diet_deg_results2$avg_log2FC) > 1, ]
head(significant_genes_glia_diet2)
dim(glia_diet_deg_results_sorted2)
dim(significant_genes_glia_diet2)
print(significant_genes_glia_diet2)
```
```{r}
glia_diet_deg_results2$log2FoldChange <- glia_diet_deg_results2$avg_log2FC
glia_diet_deg_results2$negLog10PValue <- -log10(glia_diet_deg_results2$p_val_adj)

# Remove NA values if any (they can cause plotting issues)
glia_diet_deg_results2 <- na.omit(glia_diet_deg_results2)
```

```{r}
# Create the volcano plot
EnhancedVolcano(glia_diet_deg_results2,
    lab = rownames(glia_diet_deg_results2),
    x = 'log2FoldChange',
    y = 'p_val_adj',                        
    pCutoff = 0.05, )

EnhancedVolcano(glia_diet_deg_results2,
    lab = rownames(glia_diet_deg_results2),
    x = 'log2FoldChange',
    y = 'p_val_adj',                        
    pCutoff = 0.05,
    xlim = c(-5, 5),                          
    ylim = c(0, 10), )
```

##these next two chunks separate neurons into glp1r and gfral, but we can also just use the "neurons" object and run the same DEG stuff 
##filter neurons to only Glp1r and not gfral  
```{r}
threshold <- 0
# Identify cells where at least one of the genes exceeds the threshold (| = or)
Glp1r_neurons <- rownames(gene_expression)[gene_expression$Glp1r > threshold & gene_expression$Gfral == 0]
filtered_Glp1r_neurons <- neurons[, Glp1r_neurons]
length(Cells(filtered_Glp1r_neurons)) 
saveRDS(filtered_Glp1r_neurons, "filtered_Glp1r_neurons.rds")
```

##filter neurons to only gfral and not glp1r
```{r}
threshold <- 0
# Identify cells where at least one of the genes exceeds the threshold (Gfral only)
Gfral_neurons <- rownames(gene_expression)[gene_expression$Glp1r == 0 & gene_expression$Gfral > threshold]
filtered_Gfral_neurons <- neurons[, Gfral_neurons]
length(Cells(filtered_Gfral_neurons)) 
saveRDS(filtered_Gfral_neurons, "filtered_Gfral_neurons.rds")
```
