---
title: "Project Code Final"
output: html_notebook
---

```{r}
# Load required packges:
suppressPackageStartupMessages({
library(dplyr)
install.packages("Seurat")
library(Seurat)
})
```
##this is to convert h5ad to Seurat object
#```{r}
setwd("~/Users/adriana/Documents/Stanford_Classes/First Year/winter 2026/stemrem 205/stemrem")
```

### rahcel & perla cleaned up this code so that it started with what i want 

#```{r}
metadata.DVC <- read.table("metadata_all_DVC.txt", sep = "\t", header = TRUE)
metadata.DVC <- metadata.DVC[-1,]
head(metadata.DVC)
```

##make and save to compute (saved to mac do not redo this)
#```{r}
write.csv(metadata.DVC, file = "metadata.DVC.csv", row.names = FALSE)
```


##after perla gave me terminal code for brew (do not redo this)

#```{r}
install.packages("SeuratDisk")
library(SeuratDisk)
```
#```{r}
R.version.string
packageVersion("Seurat")
install.packages("remotes")
remotes::install_github("mojaveazure/seurat-disk")
```
#```{r}
library(SeuratDisk)
install.packages("hdf5r")
```

###using perla code to convert h5ad (do not redo this)

#```{r}
Convert("SG_publication_allmice_DVC.h5ad", dest = "h5seurat", overwrite = TRUE) # will create this: your_file.h5seurat 
h5ad.as.seuratobj <- LoadH5Seurat("SG_publication_allmice_DVC.h5seurat") # will load as a seurat object and you can check it to verify, make sure you use head()
head(h5ad.as.seuratobj)
```
#```{r}
library(SeuratDisk)
SeuratDisk:::Inspect("SG_publication_allmice_DVC.h5ad")
```

##this chunk seemed to work, i see a large seurat in data (rds file is saved to mac, so we are done here)
#```{r}
library(zellkonverter)
library(Seurat)

sce <- readH5AD("SG_publication_allmice_DVC.h5ad")

allmice.h5ad.as.seurat_obj <- as.Seurat(sce, counts = "X", data = NULL)
#this step saves the seurat object as rds to my latpop 
saveRDS(allmice.h5ad.as.seurat_obj, "/Users/adriana/Documents/Stanford Classes/First Year/winter 2026/stemrem 205/stemrem1/stemrem/allmice.rds")
#```

##START HERE FOR PROJECT

##rachel helped looking for low quality: mt 
## the chunk below works to reload after deleted allmice.data from environemnt 
```{r}
library(Seurat)
allmice.data <- readRDS("/Users/adriana/Documents/Stanford Classes/First Year/winter 2026/stemrem 205/stemrem1/stemrem/allmice.rds")
head(allmice.data)
metadata <- read.csv('/Users/adriana/Documents/Stanford Classes/First Year/winter 2026/stemrem 205/stemrem1/stemrem/metadata.DVC.csv')
head(metadata)
```

##Quality check for mitchondrial genes
```{r}
allmice.data[["percent.mt"]] <- PercentageFeatureSet(allmice.data, pattern = "^MT-")
table(allmice.data$percent.mt)
allmice.data[["percent.mt"]] <- PercentageFeatureSet(allmice.data, pattern = "^(mt-|MT-)")
table(allmice.data$percent.mt)
```


#very few genes to mt 
```{r}
VlnPlot(allmice.data, features = c( "percent.mt"), ncol = 1)
```
## looks like the authors already did quality control 

##Confirming that this dataset has the genes we care about (Glp1r and Gfral)

```{r}
gene_names <- c("Glp1r", "Gfral")
present_genes <- gene_names[gene_names %in% rownames(allmice.data)]
print(present_genes)
```

```{r}
gene_expression <- FetchData(allmice.data, vars = c("Glp1r", "Gfral"))
head(gene_expression)
```

##i made an rds file with all the mouse stuff and sorted for glp1r and gfral.
##this output is GLP1 and GRFAL positive CELLS (not neurons, so any cells with at least one glp1r or gfral)
#defining threshold as greater than 0
```{r}
threshold <- 0
# Identify cells where at least one of the genes exceeds the threshold
cells_to_keep <- rownames(gene_expression)[gene_expression$Glp1r > threshold | gene_expression$Gfral > threshold]
filtered_Glp1r_Gfral <- allmice.data[, cells_to_keep]
length(Cells(filtered_Glp1r_Gfral)) 
saveRDS(filtered_Glp1r_Gfral, "filtered_Glp1r_Gfral.rds")
```

```{r}
library(dplyr)
```

#normalize my subset
```{r}
filtered_Glp1r_Gfral <- NormalizeData(filtered_Glp1r_Gfral)
```

#look for variable genes in this subset
```{r}
# identify variable genes in the group (this is not separating them)
#genes with the most variance in the dataset 
filtered_Glp1r_Gfral <- FindVariableFeatures(filtered_Glp1r_Gfral)
```

```{r}
receptor_expression <- FetchData(filtered_Glp1r_Gfral, vars = c("Glp1r", "Gfral"))
head(receptor_expression)
```

#top 10 
```{r}
top10_filtered_Glp1r_Gfral <- head(VariableFeatures(filtered_Glp1r_Gfral), 10)
top10_filtered_Glp1r_Gfral
```

# plot variablility distribution
```{r}
plot1 <- VariableFeaturePlot(filtered_Glp1r_Gfral)
plot_Glp1r_Gfral <- LabelPoints(plot = plot1, points = top10_filtered_Glp1r_Gfral, repel = TRUE)
plot_Glp1r_Gfral
```

# Z-score scale (genes)
```{r}
Glp1r.Gfral.genes <- rownames(filtered_Glp1r_Gfral)
filtered_Glp1r_Gfral <- ScaleData(filtered_Glp1r_Gfral, features = Glp1r.Gfral.genes)
```

```{r}
head(filtered_Glp1r_Gfral@assays$originalexp$counts)
head(filtered_Glp1r_Gfral@meta.data)
```
# PCA, default subsets to highly variable features
#output shows positive and negative genes
```{r}
filtered_Glp1r_Gfral <- RunPCA(filtered_Glp1r_Gfral)
```

# A visualization that is useful for determining how many PCs to include
# pick a dot for number of PCAs
```{r}
ElbowPlot(filtered_Glp1r_Gfral)
```

# computing the neighborhood graph and find clusters
# tried a bunch of different dims and res for this and this is what i like best
```{r}
filtered_Glp1r_Gfral <- FindNeighbors(filtered_Glp1r_Gfral, dims = 1:6)
filtered_Glp1r_Gfral <- FindClusters(filtered_Glp1r_Gfral, resolution = 0.15)
```

#Run UMAP
```{r}
filtered_Glp1r_Gfral <- RunUMAP(filtered_Glp1r_Gfral, dims = 1:6)
```

# plot UMAP
```{r}
DimPlot(filtered_Glp1r_Gfral, reduction = "umap")
```

## the next few steps are interrogating the clusters
```{r}
cluster_membership <- filtered_Glp1r_Gfral$seurat_clusters
table(cluster_membership)
```

### we use findallmarkers and find DIFFERENCES between clusters
##FindAllMarkersidentifies markers for all clusters in a Seurat object relative to all other clusters, differential expression analysis across all clusters 
##output is a data frame that includes a row for each gene, annotated by the cluster where it is significantly expressed
```{r}
cluster_markers <- FindAllMarkers(filtered_Glp1r_Gfral)
# View top markers for each cluster
head(cluster_markers)
```

##significant cluster markers
```{r}
significant_markers <- cluster_markers %>%
  filter(p_val_adj < 0.05 & abs(avg_log2FC) > 1) 
head(significant_markers)
dim(significant_markers)
```

##create a list of genes for each cluster
```{r}
library(dplyr)

genes_by_cluster <- significant_markers %>%
  group_by(cluster) %>%
  summarize(genes = list(gene))  

print(genes_by_cluster)
```

### look at top different genes by cluster
```{r}

#cluster0
sorted_cluster0_genes <- significant_markers %>%
  filter(cluster == "0") %>%
  arrange(desc(avg_log2FC)) %>%
  select(gene, avg_log2FC)


# View the genes for cluster0
print(sorted_cluster0_genes)

#cluster 1
sorted_cluster1_genes <- significant_markers %>%
  filter(cluster == "1") %>%
  arrange(desc(avg_log2FC)) %>%
  select(gene, avg_log2FC)

print(sorted_cluster1_genes)

#cluster 2
sorted_cluster2_genes <- significant_markers %>%
  filter(cluster == "2") %>%
  arrange(desc(avg_log2FC)) %>%
  select(gene, avg_log2FC)


print(sorted_cluster2_genes)

#cluster3
sorted_cluster3_genes <- significant_markers %>%
  filter(cluster == "3") %>%
  arrange(desc(avg_log2FC)) %>%
  select(gene, avg_log2FC)

print(sorted_cluster3_genes)

#cluster4
sorted_cluster4_genes <- significant_markers %>%
  filter(cluster == "4") %>%
  arrange(desc(avg_log2FC)) %>%
  select(gene, avg_log2FC)

print(sorted_cluster4_genes)

#cluster5
sorted_cluster5_genes <- significant_markers %>%
  filter(cluster == "5") %>%
  arrange(desc(avg_log2FC)) %>%
  select(gene, avg_log2FC)

print(sorted_cluster5_genes)

```

###I want to look at top genes for every cluster, trying to identify identity overlap !!!
##trying to figure out top genes for each cluster, not just the different ones

#0.15 res has 6 clusters (I LIKE THIS BEST)
#0.17 had 7 clusters

```{r}
library(Seurat)
library(dplyr)

# Get the average expression per cluster
average_expression <- AverageExpression(filtered_Glp1r_Gfral, return.seurat = FALSE)

# Convert the average expression to a data frame
avg_expr_df <- as.data.frame(average_expression$originalexp)

# Create an empty list to hold top genes for each cluster
top_genes_list <- list()

# Specify the number of top genes to retrieve
top_n <- 10

# Loop through each column (cluster) to find top genes
for (cluster in colnames(avg_expr_df)) {
  # Get the top N genes for the cluster
  top_genes <- avg_expr_df %>%
    arrange(desc(!!sym(cluster))) %>%  # Sort the genes by their average expression
    head(top_n) %>%                    # Select the top N genes
    rownames()                         # Extract gene names
  top_genes_list[[cluster]] <- top_genes  # Store results in the list
}

# Combine the results into a data frame 
top_genes_df <- do.call(rbind, lapply(names(top_genes_list), function(cluster_name) {
  data.frame(Cluster = cluster_name, Gene = top_genes_list[[cluster_name]], stringsAsFactors = FALSE)
}))

print(top_genes_df)
```

## separating each column is one cluster !!!
## this gives us gene is row and expr for each cluster
```{r}
library(Seurat)
library(dplyr)
library(tidyr)
library(tibble)  # For rownames_to_column()

# Get the average expression per cluster
average_expression <- AverageExpression(filtered_Glp1r_Gfral, return.seurat = FALSE)

# Convert the average expression to a data frame
avg_expr_df <- as.data.frame(average_expression$originalexp)

# Check the dim and structure of the average expression data frame
print(dim(avg_expr_df))  # Expect genes x clusters
print(head(avg_expr_df))  # Preview the first few rows

# Create an empty list to hold top genes for each cluster
top_genes_list <- list()

# Specify the number of top genes to retrieve
top_n <- 10

# Loop through each column (cluster) to find top genes along with their expression levels
for (cluster in colnames(avg_expr_df)) {
  # Get the top N genes for the cluster
  top_genes <- avg_expr_df %>%
    arrange(desc(!!sym(cluster))) %>%     # Sort the genes by their average expression
    head(top_n) %>%                       # Select the top N genes
    rownames_to_column(var = "Gene") %>%  # Convert row names to a column called "Gene"
    mutate(Cluster = cluster,              # Add a column for the cluster
           Expression = !!sym(cluster))    # Capture the corresponding expression value
  
  # Store results in the list
  top_genes_list[[cluster]] <- top_genes  
}

# Combine the results into a single data frame
top_genes_df <- bind_rows(top_genes_list)

# Pivot the data frame to have genes as rows and clusters as columns
top_genes_wide <- top_genes_df %>%
  select(Gene, Cluster, Expression) %>%  # Ensure the relevant columns are selected
  pivot_wider(names_from = Cluster, values_from = Expression, values_fill = 0)  # Convert to wide format

# View the reshaped data frame
print(top_genes_wide)
```

##this orders top genes by a specified cluster , so i am looking for the top genes in each cluster (if i want to validate the clusters the original paper created)
```{r}
#ordered by 0
ordered_genes_cluster0 <- top_genes_wide %>%
  arrange(desc(g0))  
print(ordered_genes_cluster0)

#ordered by cluster 1
ordered_genes_cluster1 <- top_genes_wide %>%
  arrange(desc(g1))  
print(ordered_genes_cluster1)

#ordered by cluster 2
ordered_genes_cluster2 <- top_genes_wide %>%
  arrange(desc(g2))  
print(ordered_genes_cluster2)

#ordered by cluster 3
ordered_genes_cluster3 <- top_genes_wide %>%
  arrange(desc(g3))  
print(ordered_genes_cluster3)

#ordered by cluster 4
ordered_genes_cluster4 <- top_genes_wide %>%
  arrange(desc(g4))  
print(ordered_genes_cluster4)

#ordered by cluster 5
ordered_genes_cluster5 <- top_genes_wide %>%
  arrange(desc(g5))  
print(ordered_genes_cluster5)

```


##trying to see which "identity_layer1" most common in cluster (identity layers assigned by original paper to the clusters they made, so i am seeing how the clusters in my dataset with only glp1r+ and gfral+ compare to their clusters of the whole DVC)

```{r}
data <- filtered_Glp1r_Gfral@meta.data %>%
  select(seurat_clusters, identity_layer1)  

# Find the most frequent identity per cluster
result <- data %>%
  group_by(seurat_clusters) %>%
  summarise(most_frequent_identity1 = names(which.max(table(identity_layer1))))

# View the result
print(result)
```

##trying identity_layer2 - NO REPEATS WITH 0.15 res, so we like 0.15 res better than 0.2 , 6 total clusters and label by identity layer 2, but astrocytes get messed up. trying other res . 0.17 has repeats so stick with 0.15!
```{r}
data2 <- filtered_Glp1r_Gfral@meta.data %>%
  select(seurat_clusters, identity_layer2)  # Make sure 'identity' is the correct name for the identity column

# Find the most frequent identity per cluster
result2 <- data2 %>%
  group_by(seurat_clusters) %>%
  summarise(most_frequent_identity2 = names(which.max(table(identity_layer2))))

# View the result
print(result2)

#how much of each cluster is id2
result2 <- data2 %>%
  group_by(seurat_clusters, identity_layer2) %>%
  summarise(count = n()) %>%
  ungroup() %>%
  group_by(seurat_clusters) %>%
  mutate(proportion = count / sum(count)) %>%
  select(seurat_clusters, identity_layer2, proportion) %>%
  arrange(seurat_clusters, desc(proportion))

# View the result
print(result2)
```


##trying identity_layer3
```{r}
data3 <- filtered_Glp1r_Gfral@meta.data %>%
  select(seurat_clusters, identity_layer3)  # Make sure 'identity' is the correct name for the identity column

# Find the most frequent identity per cluster
result3 <- data3 %>%
  group_by(seurat_clusters) %>%
  summarise(most_frequent_identity3 = names(which.max(table(identity_layer3))))

##how much of each cluster has this identity3
result3 <- data3 %>%
  group_by(seurat_clusters, identity_layer3) %>%
  summarise(count = n()) %>%
  ungroup() %>%
  group_by(seurat_clusters) %>%
  mutate(proportion = count / sum(count)) %>%
  select(seurat_clusters, identity_layer3, proportion) %>%
  arrange(seurat_clusters, desc(proportion))

# View the result
print(result3)
```

## i like it better with 0.15 res ! keeping this
##rename clusters by identity 

```{r}
# 0 = mixed_neurons (because it has a lot of 1. 2 and 3) (layer 3 but more specific than layer 2 i think)
# 1 = Sall3_class (layer 2, more accurate, 70%) (are these progenitors maybe?)
# 2 = M0 (M0 good, 98%) (monoamine)
# 3 = Vglut1 (good, 85%) (excitatory)
# 4 = astrocyte (this is mixed but lets keep it?? figure out bio relevance- golli neurons express glia genes!!!)
# 5 = oligodendrocyte (98%)
# 6 = 

new.cluster.ids <- c("Mixed Neurons", 'Sall3 Neurons', "Monoamine Neurons", 
                     'Excitatory Neurons', 'Astrocyte',
                     'Oligodendrocyte'
                     )
names(new.cluster.ids) <- levels(filtered_Glp1r_Gfral)
filtered_Glp1r_Gfral <- RenameIdents(filtered_Glp1r_Gfral, new.cluster.ids)
```

```{r}
# plot umap with renamed clusters
library(ggplot2)
DimPlot(filtered_Glp1r_Gfral, reduction = "umap", label = TRUE, pt.size = 0.3) +
  theme(
    legend.position = c(0.78, 0.5),  # further right & lower
   plot.margin = margin(t = 5, r = 40, b = 5, l = 5)  # increase right margin
  ) +
  coord_cartesian(clip = "off")
```

##which clusters have highest expression of our target genes? 
```{r}
genes_to_plot = c( 'Glp1r', 'Gfral')
#FeaturePlot(filtered_Glp1r, reduction = "umap", features = genes_to_plot)
#FeaturePlot(filtered_Gfral, reduction = "umap", features = genes_to_plot)
FeaturePlot(filtered_Glp1r_Gfral, reduction = "umap", features = "Glp1r")
FeaturePlot(filtered_Glp1r_Gfral, reduction = "umap", features = "Gfral")
```

##NOW MOVING ON TO DEGS between Glp1r+ cells vs Gfral+ cells 
#identifying cells to pick
```{r}
glp1r_cells <- WhichCells(filtered_Glp1r_Gfral, expression = Glp1r > 0)  # Adjust the threshold based on your needs
gfral_cells <- WhichCells(filtered_Glp1r_Gfral, expression = Gfral > 0) 
```

#this makes a column called cell_type and labels cells as both, glp1r, or gfral so we can compare only glp1r and gfral
##FindMarkers compares our two groups and the output is DEGs between the two groups
```{r}
filtered_Glp1r_Gfral <- AddMetaData(filtered_Glp1r_Gfral, 
                                     metadata = ifelse(Cells(filtered_Glp1r_Gfral) %in% glp1r_cells & 
                                                       Cells(filtered_Glp1r_Gfral) %in% gfral_cells, 
                                                       "Both", 
                                                       ifelse(Cells(filtered_Glp1r_Gfral) %in% glp1r_cells, 
                                                              "Glp1r", 
                                                              "Gfral")), 
                                     col.name = "cell_type")

# Identify differentially expressed genes between the two cell types of interest, this selected the cells we want and not BOTH
deg_results <- FindMarkers(filtered_Glp1r_Gfral, ident.1 = "Glp1r", ident.2 = "Gfral", group.by = "cell_type")

# View the results
## this provides q value by BH
head(deg_results)
```

#sort the degs 
```{r}
deg_results_sorted <- deg_results[order(deg_results$avg_log2FC), ]
head(deg_results_sorted)
significant_genes <- deg_results[deg_results$p_val_adj < 0.05 & abs(deg_results$avg_log2FC) > 1, ]
head(significant_genes)
dim(deg_results_sorted)
dim(significant_genes)
print(significant_genes)
```

```{r}
# Create a new data frame for plotting
library(ggplot2)

# Add a column for -log10(p_val_adj) for plotting
deg_results$log2FoldChange <- deg_results$avg_log2FC
deg_results$negLog10PValue <- -log10(deg_results$p_val_adj)

# Remove NA values if any (they can cause plotting issues)
deg_results <- na.omit(deg_results)
```

##volcano with enhanced volcano
```{r}
BiocManager::install("EnhancedVolcano", force = TRUE)
library(EnhancedVolcano)
```

##this is showing us DEGs, first plot shows glp1r and gfral then we narrow it down to see the other genes 
```{r}
# Create the volcano plot
EnhancedVolcano(deg_results,
    lab = rownames(deg_results),
    x = 'log2FoldChange',
    y = 'p_val_adj',                        
    pCutoff = 0.05, )

EnhancedVolcano(deg_results,
    lab = rownames(deg_results),
    x = 'log2FoldChange',
    y = 'p_val_adj',                        
    pCutoff = 0.05,
    xlim = c(-5, 5),                          
    ylim = c(0, 10), )
```

### NOW WE ARE SUBSETTING TO ONLY NEURONS TO DO DE ANALYSIS 

#this subsets to only neurons 
```{r}
neurons <- subset(filtered_Glp1r_Gfral, subset = identity_layer1 == "neuron")
head(neurons)
saveRDS(neurons, "filtered_Glp1r_Gfral_neurons.rds")
```

##looking at degs between glp1r and gfral neurons !
# Identify differentially expressed genes between the two cell types of interest 
```{r}
neuron_deg_results <- FindMarkers(neurons, ident.1 = "Glp1r", ident.2 = "Gfral", group.by = "cell_type")

## this provides q value by BH
head(neuron_deg_results)
```

```{r}
neuron_deg_results_sorted <- neuron_deg_results[order(neuron_deg_results$avg_log2FC), ]
head(neuron_deg_results_sorted)
significant_genes_neuron <- neuron_deg_results[neuron_deg_results$p_val_adj < 0.05 & abs(neuron_deg_results$avg_log2FC) > 1, ]
head(significant_genes_neuron)
dim(neuron_deg_results_sorted)
dim(significant_genes_neuron)
print(significant_genes_neuron)
```

```{r}
neuron_deg_results$log2FoldChange <- neuron_deg_results$avg_log2FC
neuron_deg_results$negLog10PValue <- -log10(neuron_deg_results$p_val_adj)

# Remove NA values if any (they can cause plotting issues)
neuron_deg_results <- na.omit(neuron_deg_results)
```

#Neuron DEG volcano plot
```{r}
EnhancedVolcano(neuron_deg_results,
    lab = rownames(neuron_deg_results),
    x = 'log2FoldChange',
    y = 'p_val_adj',                        
    pCutoff = 0.05, )

EnhancedVolcano(neuron_deg_results,
    lab = rownames(neuron_deg_results),
    x = 'log2FoldChange',
    y = 'p_val_adj',                        
    pCutoff = 0.05,
    xlim = c(-5, 5),                          
    ylim = c(0, 10), )
```


## repeat to compare DEGs in Glp1r+ vs Gfral+ glia

### Now we are subsetting to ONLY GLIA. repeat the above DEG analysis from the whole cell population with only GLIA with our receptors. this is GLIA only 

```{r}
distinct_identity <- unique(filtered_Glp1r_Gfral$identity_layer1)
print(distinct_identity)
```
#SUBSET TO ONLY GLIA 
```{r}
glia <- subset(filtered_Glp1r_Gfral, subset = identity_layer1 == "glial")
head(glia)
saveRDS(glia, "filtered_Glp1r_Gfral_glia.rds")
dim(glia)
```
# Identify differentially expressed genes in glia between the two cell types of interest (receptors)
```{r}
glia_deg_results <- FindMarkers(glia, ident.1 = "Glp1r", ident.2 = "Gfral", group.by = "cell_type")

## this provides q value by BH
head(glia_deg_results)
```

```{r}
glia_deg_results_sorted <- glia_deg_results[order(glia_deg_results$avg_log2FC), ]
head(glia_deg_results_sorted)
significant_genes_glia <- glia_deg_results[glia_deg_results$p_val_adj < 0.05 & abs(glia_deg_results$avg_log2FC) > 1, ]
head(significant_genes_glia)
dim(glia_deg_results_sorted)
dim(significant_genes_glia)
print(significant_genes_glia)
```

```{r}
glia_deg_results$log2FoldChange <- glia_deg_results$avg_log2FC
glia_deg_results$negLog10PValue <- -log10(glia_deg_results$p_val_adj)

# Remove NA values if any (they can cause plotting issues)
glia_deg_results <- na.omit(glia_deg_results)
```
#Glia volcano plot 
```{r}
EnhancedVolcano(glia_deg_results,
    lab = rownames(glia_deg_results),
    x = 'log2FoldChange',
    y = 'p_val_adj',                        
    pCutoff = 0.05, )

EnhancedVolcano(glia_deg_results,
    lab = rownames(glia_deg_results),
    x = 'log2FoldChange',
    y = 'p_val_adj',                        
    pCutoff = 0.05,
    xlim = c(-5, 5),                          
    ylim = c(0, 10), )
```


##if we want to look at nutritional states, need to be able to separate feeding conditions 
##Introducing a new label into the metadata so we can look at feeding conditions
```{r}
distinct_values <- unique(filtered_Glp1r_Gfral$orig.ident)
print(distinct_values)
```
#creating a new column with the name diet_category like we did with cell_type (THIS IS NOT SPECIFIC TO NEURONS!)
```{r}
filtered_Glp1r_Gfral <- AddMetaData(
  object = filtered_Glp1r_Gfral,
  metadata = case_when(
    filtered_Glp1r_Gfral$orig.ident == "adlib" ~ "Ad libitum",
    filtered_Glp1r_Gfral$orig.ident == "fast" ~ "Fasted",
    filtered_Glp1r_Gfral$orig.ident == "refed" ~ "Refed",  
    TRUE ~ "Other"  # Fallback for any other conditions not specifically captured
  ),
  col.name = "diet_category"
)

head(filtered_Glp1r_Gfral)
# Create a frequency table for the 'diet' column
diet_counts <- table(filtered_Glp1r_Gfral$diet_category)

# Print the counts
print(diet_counts)
```

## NOTE: in the other R Notebook we did many analyses related to diet. HOWEVER, there was a ton of skew related to sex genes, and the authors of the original dataset did not separate by sex. i think there may be some skew in the samples, not biological relevance, so we have removed these analysis from our presentation. They can still be found in the original "project_clean.Rmd" notebook :) 

##to keep with the theme, we instead are comparing how Glp1r+ versus Gfral+ cells respond to different states, instead of comparing the feeding states 

## looking how glp1r+ vs gfral+ respond to diet states differently? 
###filtering neurons to nutritional states to compare glp1r vs gfral in fasted and then doing the same with ad lib and refed 

```{r}
neurons_adlib <- subset(filtered_Glp1r_Gfral, subset = diet_category == "Ad libitum")
head(neurons_adlib)
neurons_fasted <- subset(filtered_Glp1r_Gfral, subset = diet_category == "Fasted")
head(neurons_fasted)
neurons_refed <- subset(filtered_Glp1r_Gfral, subset = diet_category == "Refed")
head(neurons_refed)
```

##Start with ad lib condition! 
### there is no difference in adlib! 

```{r}
neurons_adlib_deg_results <- FindMarkers(neurons_adlib, ident.1 = "Glp1r", ident.2 = "Gfral", group.by = "cell_type")

# View the results
## this provides q value by BH
head(neurons_adlib_deg_results)
```
```{r}
neurons_adlib_deg_results <- neurons_adlib_deg_results[order(neurons_adlib_deg_results$avg_log2FC), ]
head(neurons_adlib_deg_results)
significant_genes_neuron_adlib <- neurons_adlib_deg_results[neurons_adlib_deg_results$p_val_adj < 0.05 & abs(neurons_adlib_deg_results$avg_log2FC) > 1, ]
head(significant_genes_neuron_adlib)
dim(neurons_adlib_deg_results)
dim(significant_genes_neuron_adlib)
print(significant_genes_neuron_adlib)
```
```{r}
neurons_adlib_deg_results$log2FoldChange <- neurons_adlib_deg_results$avg_log2FC
neurons_adlib_deg_results$negLog10PValue <- -log10(neurons_adlib_deg_results$p_val_adj)

# Remove NA values if any (they can cause plotting issues)
neurons_adlib_deg_results <- na.omit(neurons_adlib_deg_results)
```


```{r}
# Create the volcano plot
EnhancedVolcano(neurons_adlib_deg_results,
    lab = rownames(neurons_adlib_deg_results),
    x = 'log2FoldChange',
    y = 'p_val_adj',                        
    pCutoff = 0.05, )

EnhancedVolcano(neurons_adlib_deg_results,
    lab = rownames(neurons_adlib_deg_results),
    x = 'log2FoldChange',
    y = 'p_val_adj',                        
    pCutoff = 0.05,
    xlim = c(-5, 5),                          
    ylim = c(0, 10), )
```

## compare glp1r+ vs gfral+ in FASTED states 
```{r}
neurons_fasted_deg_results <- FindMarkers(neurons_fasted, ident.1 = "Glp1r", ident.2 = "Gfral", group.by = "cell_type")

## this provides q value by BH
head(neurons_fasted_deg_results)
```

```{r}
neurons_fasted_deg_results <- neurons_fasted_deg_results[order(neurons_fasted_deg_results$avg_log2FC), ]
head(neurons_fasted_deg_results)
significant_genes_neuron_fasted <- neurons_fasted_deg_results[neurons_fasted_deg_results$p_val_adj < 0.05 & abs(neurons_fasted_deg_results$avg_log2FC) > 1, ]
head(significant_genes_neuron_fasted)
dim(neurons_fasted_deg_results)
dim(significant_genes_neuron_fasted)
print(significant_genes_neuron_fasted)
```
```{r}
neurons_fasted_deg_results$log2FoldChange <- neurons_fasted_deg_results$avg_log2FC
neurons_fasted_deg_results$negLog10PValue <- -log10(neurons_fasted_deg_results$p_val_adj)

# Remove NA values if any (they can cause plotting issues)
neurons_fasted_deg_results <- na.omit(neurons_fasted_deg_results)
```

```{r}
# Create the volcano plot
EnhancedVolcano(neurons_fasted_deg_results,
    lab = rownames(neurons_fasted_deg_results),
    x = 'log2FoldChange',
    y = 'p_val_adj',                        
    pCutoff = 0.05, )

EnhancedVolcano(neurons_fasted_deg_results,
    lab = rownames(neurons_fasted_deg_results),
    x = 'log2FoldChange',
    y = 'p_val_adj',                        
    pCutoff = 0.05,
    xlim = c(-5, 5),                          
    ylim = c(0, 10), )
```

### now looking at refed conditions glp1r+ vs gfral+ neurons

```{r}
neurons_refed_deg_results <- FindMarkers(neurons_refed, ident.1 = "Glp1r", ident.2 = "Gfral", group.by = "cell_type")

# View the results
## this provides q value by BH
head(neurons_refed_deg_results)
```

```{r}
neurons_refed_deg_results <- neurons_refed_deg_results[order(neurons_refed_deg_results$avg_log2FC), ]
head(neurons_refed_deg_results)
significant_genes_neuron_refed <- neurons_refed_deg_results[neurons_refed_deg_results$p_val_adj < 0.05 & abs(neurons_refed_deg_results$avg_log2FC) > 1, ]
head(significant_genes_neuron_refed)
dim(neurons_refed_deg_results)
dim(significant_genes_neuron_refed)
print(significant_genes_neuron_refed)
```

```{r}
neurons_refed_deg_results$log2FoldChange <- neurons_refed_deg_results$avg_log2FC
neurons_refed_deg_results$negLog10PValue <- -log10(neurons_refed_deg_results$p_val_adj)

# Remove NA values if any (they can cause plotting issues)
neurons_refed_deg_results <- na.omit(neurons_refed_deg_results)
```

```{r}
# Create the volcano plot
EnhancedVolcano(neurons_refed_deg_results,
    lab = rownames(neurons_refed_deg_results),
    x = 'log2FoldChange',
    y = 'p_val_adj',                        
    pCutoff = 0.05, )

EnhancedVolcano(neurons_refed_deg_results,
    lab = rownames(neurons_refed_deg_results),
    x = 'log2FoldChange',
    y = 'p_val_adj',                        
    pCutoff = 0.05,
    xlim = c(-5, 5),                          
    ylim = c(0, 10), )
```

