---
title: "project_clean"
output: html_notebook
---


```{r}
# Load required packges:
suppressPackageStartupMessages({
library(dplyr)
install.packages("Seurat")
library(Seurat)
install.packages("DESeq2")
})
```

```{r}
setwd("~/Users/adriana/Documents/Stanford_Classes/First Year/winter 2026/stemrem 205/stemrem")
```

### rahcel & perla cleaned up this code so that it started with what i want 

```{r}
metadata.DVC <- read.table("metadata_all_DVC.txt", sep = "\t", header = TRUE)
metadata.DVC <- metadata.DVC[-1,]
head(metadata.DVC)
```

##make and save to compute (saved to mac do not redo this)
```{r}
write.csv(metadata.DVC, file = "metadata.DVC.csv", row.names = FALSE)
```


##after perla gave me terminal code for brew (do not redo this)
```{r}
install.packages("SeuratDisk")
library(SeuratDisk)
```
```{r}
R.version.string
packageVersion("Seurat")
install.packages("remotes")
remotes::install_github("mojaveazure/seurat-disk")
```
```{r}
library(SeuratDisk)
install.packages("hdf5r")
```

###using perla code to convert h5ad (do not redo this)

```{r}
Convert("SG_publication_allmice_DVC.h5ad", dest = "h5seurat", overwrite = TRUE) # will create this: your_file.h5seurat 
h5ad.as.seuratobj <- LoadH5Seurat("SG_publication_allmice_DVC.h5seurat") # will load as a seurat object and you can check it to verify, make sure you use head()
head(h5ad.as.seuratobj)
```
```{r}
library(SeuratDisk)
SeuratDisk:::Inspect("SG_publication_allmice_DVC.h5ad")
```

##this chunk seemed to work, i see a large seurat in data (rds file is saved to mac, so we are done here)
```{r}
library(zellkonverter)
library(Seurat)

sce <- readH5AD("SG_publication_allmice_DVC.h5ad")

allmice.h5ad.as.seurat_obj <- as.Seurat(sce, counts = "X", data = NULL)
#this step saves the seurat object as rds to my latpop 
saveRDS(allmice.h5ad.as.seurat_obj, "/Users/adriana/Documents/Stanford Classes/First Year/winter 2026/stemrem 205/stemrem1/stemrem/allmice.rds")
```

#### START HERE FOR PROJECT

##rachel helped looking for low quality: mt 
## the chunk below works to reload after deleted allmice.data from environemnt 
```{r}
library(Seurat)
allmice.data <- readRDS("/Users/adriana/Documents/Stanford Classes/First Year/winter 2026/stemrem 205/stemrem1/stemrem/allmice.rds")
head(allmice.data)
metadata <- read.csv('metadata.DVC.csv')
head(metadata)
```
```{r}
allmice.data[["percent.mt"]] <- PercentageFeatureSet(allmice.data, pattern = "^MT-")
table(allmice.data$percent.mt)
```
```{r}
allmice.data[["percent.mt"]] <- PercentageFeatureSet(allmice.data, pattern = "^(mt-|MT-)")
table(allmice.data$percent.mt)
```

#very few genes to mt 
```{r}
VlnPlot(allmice.data, features = c( "percent.mt"), ncol = 1)
```


```{r}
features <- rownames(allmice.data)
head(features)
```


#looking to make sure these exist 
```{r}
gene_names <- c("Glp1r", "Gfral")
present_genes <- gene_names[gene_names %in% rownames(allmice.data)]
print(present_genes)
```

```{r}
gene_expression <- FetchData(allmice.data, vars = c("Glp1r", "Gfral"))
head(gene_expression)
```

##i made an rds file with all the mouse stuff and sorted for glp1r and gfral. but i did not sort mitochondria or features so may need to redo!!! (its we checked mitchondria and it seems like the authors already sorted that )
##this output is GLP1 and GRFAL positive CELLS (not neurons, so cells with both things)
#defining threshold as greater than 0
```{r}
threshold <- 0
# Identify cells where at least one of the genes exceeds the threshold
cells_to_keep <- rownames(gene_expression)[gene_expression$Glp1r > threshold | gene_expression$Gfral > threshold]
filtered_Glp1r_Gfral <- allmice.data[, cells_to_keep]
length(Cells(filtered_Glp1r_Gfral)) 
saveRDS(filtered_Glp1r_Gfral, "filtered_Glp1r_Gfral.rds")
```

##filter to only Glp1r and not gfral, this output is only glp1 
```{r}
threshold <- 0
# Identify cells where at least one of the genes exceeds the threshold (| = or)
Glp1r_cells_to_keep <- rownames(gene_expression)[gene_expression$Glp1r > threshold & gene_expression$Gfral == 0]
filtered_Glp1r <- allmice.data[, Glp1r_cells_to_keep]
length(Cells(filtered_Glp1r)) 
saveRDS(filtered_Glp1r, "filtered_Glp1r.rds")
```

##filter to only gfral, this output is only gfral
```{r}
threshold <- 0
# Identify cells where at least one of the genes exceeds the threshold (Gfral only)
Gfral_cells_to_keep <- rownames(gene_expression)[gene_expression$Glp1r == 0 & gene_expression$Gfral > threshold]
filtered_Gfral <- allmice.data[, Gfral_cells_to_keep]
length(Cells(filtered_Gfral)) 
saveRDS(filtered_Gfral, "filtered_Gfral.rds")
```

```{r}
library(dplyr)
```

#normal set with both and each separate
```{r}
# normalize
filtered_Glp1r_Gfral <- NormalizeData(filtered_Glp1r_Gfral)
filtered_Gfral <- NormalizeData(filtered_Gfral)
filtered_Glp1r <- NormalizeData(filtered_Glp1r)
```

#look for variable genes in combined with both glp1 and gfral 
```{r}
# identify variable genes in the group (this is not separating them)
#genes with the most variance in the dataset 
filtered_Glp1r_Gfral <- FindVariableFeatures(filtered_Glp1r_Gfral)
```
```{r}
receptor_expression <- FetchData(filtered_Glp1r_Gfral, vars = c("Glp1r", "Gfral"))
head(receptor_expression)
```

#variable genes in separate
```{r}
##identify variable genes between these 
filtered_Glp1r <- FindVariableFeatures(filtered_Glp1r)
filtered_Gfral <- FindVariableFeatures(filtered_Gfral)
```

# top 10 for each separated 
```{r}
# view top 10
top10_Glp1r <- head(VariableFeatures(filtered_Glp1r), 10)
top10_Gfral <- head(VariableFeatures(filtered_Gfral), 10)
top10_Glp1r
top10_Gfral
```

#plot each separately

```{r}
# plot variablility distribution
plot1 <- VariableFeaturePlot(filtered_Glp1r)
plot_Glp1r <- LabelPoints(plot = plot1, points = top10_Glp1r, repel = TRUE)
plot2 <- VariableFeaturePlot(filtered_Gfral)
plot_Gfral <- LabelPoints(plot = plot2, points = top10_Gfral, repel = TRUE) 
plot_Glp1r
plot_Gfral
```


```{r}
# Z-score scale (genes)
Glp1r.Gfral.genes <- rownames(filtered_Glp1r_Gfral)
Glp1r.genes <- rownames(filtered_Glp1r)
filtered_Glp1r <- ScaleData(filtered_Glp1r, features = Glp1r.genes)
Gfral.genes <- rownames(filtered_Gfral)
filtered_Gfral <- ScaleData(filtered_Gfral, features = Gfral.genes)
filtered_Glp1r_Gfral <- ScaleData(filtered_Glp1r_Gfral, features = Glp1r.Gfral.genes)
```

```{r}
head(filtered_Glp1r@assays$originalexp$counts)
head(filtered_Glp1r@meta.data)
head(filtered_Gfral@assays$originalexp$counts)
head(filtered_Gfral@meta.data)
```

# PCA, default subsets to highly variable features
#output shows positive and negative genes
```{r}
filtered_Glp1r <- RunPCA(filtered_Glp1r)
filtered_Gfral <- RunPCA(filtered_Gfral)
filtered_Glp1r_Gfral <- RunPCA(filtered_Glp1r_Gfral)
```

###how do i spread out the cells more so they arent all on top of each other like this? 
```{r}
# A visualization that is useful for determining how many PCs to include
# pick a dot for number of PCAs
ElbowPlot(filtered_Glp1r)
ElbowPlot(filtered_Gfral)
ElbowPlot(filtered_Glp1r_Gfral)
```

# computing the neighborhood graph and find clusters for each receptor type and also all my cells 
#higher resolution is more clusters
##starting with 10 
```{r}
filtered_Glp1r <- FindNeighbors(filtered_Glp1r, dims = 1:5)
filtered_Glp1r <- FindClusters(filtered_Glp1r, resolution = 0.2)
filtered_Gfral <- FindNeighbors(filtered_Gfral, dims = 1:4)
filtered_Gfral <- FindClusters(filtered_Gfral, resolution = 0.2)
filtered_Glp1r_Gfral <- FindNeighbors(filtered_Glp1r_Gfral, dims = 1:6)
filtered_Glp1r_Gfral <- FindClusters(filtered_Glp1r_Gfral, resolution = 0.2)
```

#UMAP
```{r}
filtered_Glp1r <- RunUMAP(filtered_Glp1r, dims = 1:5)
filtered_Gfral <- RunUMAP(filtered_Gfral, dims = 1:4)
filtered_Glp1r_Gfral <- RunUMAP(filtered_Glp1r_Gfral, dims = 1:6)
```

# plot UMAP
```{r}
DimPlot(filtered_Glp1r, reduction = "umap")
DimPlot(filtered_Gfral, reduction = "umap")
DimPlot(filtered_Glp1r_Gfral, reduction = "umap")
```

##looking at interesting genes on all of the umaps
```{r}
genes_to_plot = c('Phox2b', 'Glp1r', 'Gfral', 'Cck', 'Th', 'Vglut', 'Lmx1b', 'Gcg', 'Calcr', 'Gad2', 'Casr', 'Glp1', 'Gdf15' )
FeaturePlot(filtered_Glp1r, reduction = "umap", features = genes_to_plot)
FeaturePlot(filtered_Gfral, reduction = "umap", features = genes_to_plot)
FeaturePlot(filtered_Glp1r_Gfral, reduction = "umap", features = genes_to_plot)
```

## compare DEGs between glp1r and gfral cells using the combined dataset 

#identifying cells to pick
```{r}
glp1r_cells <- WhichCells(filtered_Glp1r_Gfral, expression = Glp1r > 0)  # Adjust the threshold based on your needs
gfral_cells <- WhichCells(filtered_Glp1r_Gfral, expression = Gfral > 0) 
```

#this makes a column called cell_type and labels cells as both, glp1r, or gfral 
```{r}
filtered_Glp1r_Gfral <- AddMetaData(filtered_Glp1r_Gfral, 
                                     metadata = ifelse(Cells(filtered_Glp1r_Gfral) %in% glp1r_cells & 
                                                       Cells(filtered_Glp1r_Gfral) %in% gfral_cells, 
                                                       "Both", 
                                                       ifelse(Cells(filtered_Glp1r_Gfral) %in% glp1r_cells, 
                                                              "Glp1r", 
                                                              "Gfral")), 
                                     col.name = "cell_type")

# Identify differentially expressed genes between the two cell types of interest 
deg_results <- FindMarkers(filtered_Glp1r_Gfral, ident.1 = "Glp1r", ident.2 = "Gfral", group.by = "cell_type")

# View the results
## this provides q value by BH
head(deg_results)
```

#sort the degs 
```{r}
deg_results_sorted <- deg_results[order(deg_results$avg_log2FC), ]
head(deg_results_sorted)
significant_genes <- deg_results[deg_results$p_val_adj < 0.05 & abs(deg_results$avg_log2FC) > 1, ]
head(significant_genes)
dim(deg_results_sorted)
dim(significant_genes)
print(significant_genes)
```

```{r}
# Create a new data frame for plotting
library(ggplot2)

# Add a column for -log10(p_val_adj) for plotting
deg_results$log2FoldChange <- deg_results$avg_log2FC
deg_results$negLog10PValue <- -log10(deg_results$p_val_adj)

# Remove NA values if any (they can cause plotting issues)
deg_results <- na.omit(deg_results)
```

##volcano with enhanced volcano
```{r}
BiocManager::install("EnhancedVolcano")
```

##this is showing us DEGs, first plot shows glp1r and gfral then we narrow it down to see the other genes 
```{r}
# Create the volcano plot
EnhancedVolcano(deg_results,
    lab = rownames(deg_results),
    x = 'log2FoldChange',
    y = 'p_val_adj',                        
    pCutoff = 0.05, )

EnhancedVolcano(deg_results,
    lab = rownames(deg_results),
    x = 'log2FoldChange',
    y = 'p_val_adj',                        
    pCutoff = 0.05,
    xlim = c(-5, 5),                          
    ylim = c(0, 10), )
```

### Now we are subsetting to ONLY neurons. repeat the above DEG analysis from the whole cell population with only neurons 

```{r}
neurons <- subset(filtered_Glp1r_Gfral, subset = identity_layer1 == "neuron")
head(neurons)
saveRDS(neurons, "filtered_Glp1r_Gfral_neurons.rds")
```

##looking at degs between glp1r and gfral neurons only 

```{r}
# Identify differentially expressed genes between the two cell types of interest 
neuron_deg_results <- FindMarkers(neurons, ident.1 = "Glp1r", ident.2 = "Gfral", group.by = "cell_type")

# View the results
## this provides q value by BH
head(neuron_deg_results)
```

```{r}
neuron_deg_results_sorted <- neuron_deg_results[order(neuron_deg_results$avg_log2FC), ]
head(neuron_deg_results_sorted)
significant_genes_neuron <- neuron_deg_results[neuron_deg_results$p_val_adj < 0.05 & abs(neuron_deg_results$avg_log2FC) > 1, ]
head(significant_genes_neuron)
dim(neuron_deg_results_sorted)
dim(significant_genes_neuron)
print(significant_genes_neuron)
```
```{r}
neuron_deg_results$log2FoldChange <- neuron_deg_results$avg_log2FC
neuron_deg_results$negLog10PValue <- -log10(neuron_deg_results$p_val_adj)

# Remove NA values if any (they can cause plotting issues)
neuron_deg_results <- na.omit(neuron_deg_results)
```

```{r}
# Create the volcano plot
EnhancedVolcano(neuron_deg_results,
    lab = rownames(neuron_deg_results),
    x = 'log2FoldChange',
    y = 'p_val_adj',                        
    pCutoff = 0.05, )

EnhancedVolcano(neuron_deg_results,
    lab = rownames(neuron_deg_results),
    x = 'log2FoldChange',
    y = 'p_val_adj',                        
    pCutoff = 0.05,
    xlim = c(-5, 5),                          
    ylim = c(0, 10), )
```


### now looking between adlib vs fasted (can do this with any feeding condition)

###DEGS for ad libitum vs fast in our subset !!! 
## i think this pattern can be used for like neurons and stuff too. NEED to read paper to see what is already done! but i think this way of subsetting is unique 
```{r}
distinct_values <- unique(filtered_Glp1r_Gfral$orig.ident)
print(distinct_values)
```


#creating a new column with the name diet_category like we did with cell_type 
```{r}
filtered_Glp1r_Gfral <- AddMetaData(
  object = filtered_Glp1r_Gfral,
  metadata = case_when(
    filtered_Glp1r_Gfral$orig.ident == "adlib" ~ "Ad libitum",
    filtered_Glp1r_Gfral$orig.ident == "fast" ~ "Fasted",
    filtered_Glp1r_Gfral$orig.ident == "refed" ~ "Refed",  
    TRUE ~ "Other"  # Fallback for any other conditions not specifically captured
  ),
  col.name = "diet_category"
)

head(filtered_Glp1r_Gfral)
# Create a frequency table for the 'diet' column
diet_counts <- table(filtered_Glp1r_Gfral$diet_category)

# Print the counts
print(diet_counts)
```

```{r}
# Identify differentially expressed genes between the diet types
diet_deg_results <- FindMarkers(filtered_Glp1r_Gfral, ident.1 = "Ad libitum", ident.2 = "Fasted", group.by = "diet_category")

# View the results
## this provides q value by BH
head(diet_deg_results)
```

```{r}
diet_deg_results_sorted <- diet_deg_results[order(diet_deg_results$avg_log2FC), ]
head(diet_deg_results_sorted)
significant_genes_diet <- diet_deg_results[diet_deg_results$p_val_adj < 0.05 & abs(diet_deg_results$avg_log2FC) > 1, ]
head(significant_genes_diet)
dim(diet_deg_results_sorted)
dim(significant_genes_diet)
print(significant_genes_diet)
```
```{r}
diet_deg_results$log2FoldChange <- diet_deg_results$avg_log2FC
diet_deg_results$negLog10PValue <- -log10(diet_deg_results$p_val_adj)

# Remove NA values if any (they can cause plotting issues)
diet_deg_results <- na.omit(diet_deg_results)
```

```{r}
# Create the volcano plot
EnhancedVolcano(diet_deg_results,
    lab = rownames(diet_deg_results),
    x = 'log2FoldChange',
    y = 'p_val_adj',                        
    pCutoff = 0.05, )

EnhancedVolcano(diet_deg_results,
    lab = rownames(diet_deg_results),
    x = 'log2FoldChange',
    y = 'p_val_adj',                        
    pCutoff = 0.05,
    xlim = c(-5, 5),                          
    ylim = c(0, 10), )
```

##trying to label glp1r and gfral in volcano plot for fed vs fasted, they are there but not significant 

```{r}
# Define the genes you want to check
genes_to_check <- c("Glp1r", "Gfral")

# Check if these genes are in the row names of the deg_results
gene_presence <- genes_to_check %in% rownames(diet_deg_results)

# Show which genes are present
print(data.frame(Gene = genes_to_check, Present = gene_presence))
presence_results <- data.frame(Gene = genes_to_check, Present = gene_presence)
print(presence_results)
# Get the rows corresponding to the present genes
present_genes_info <- diet_deg_results[rownames(diet_deg_results) %in% genes_to_check, ]

# Print the information for the present genes
print(present_genes_info)
```

##now looking at DEGs bw neurons and glia in my population

##neurons vs glia identity1 first using both combined 
```{r}
# Perform differential expression analysis comparing neurons and glia
deg_neuron_vs_glia <- FindMarkers(filtered_Glp1r_Gfral, ident.1 = "neuron", ident.2 = "glial", group.by = "identity_layer1")

# View the results
head(deg_neuron_vs_glia)
```
```{r}
deg_neuron_vs_glia$log2FoldChange <- deg_neuron_vs_glia$avg_log2FC
deg_neuron_vs_glia$negLog10PValue <- -log10(deg_neuron_vs_glia$p_val_adj)

# Remove NA values if any (they can cause plotting issues)
deg_neuron_vs_glia <- na.omit(deg_neuron_vs_glia)
```

```{r}
# Create the volcano plot
EnhancedVolcano(deg_neuron_vs_glia,
    lab = rownames(deg_neuron_vs_glia),
    x = 'log2FoldChange',
    y = 'p_val_adj',                        
    pCutoff = 0.05, )

EnhancedVolcano(deg_neuron_vs_glia,
    lab = rownames(deg_neuron_vs_glia),
    x = 'log2FoldChange',
    y = 'p_val_adj',                        
    pCutoff = 0.05,
    xlim = c(-5, 5),                          
    ylim = c(0, 10), )
```




##these next two chunks separate neurons into glp1r and gfral, but we can also just use the "neurons" object and run the same DEG stuff 
##filter neurons to only Glp1r and not gfral  
```{r}
threshold <- 0
# Identify cells where at least one of the genes exceeds the threshold (| = or)
Glp1r_neurons <- rownames(gene_expression)[gene_expression$Glp1r > threshold & gene_expression$Gfral == 0]
filtered_Glp1r_neurons <- neurons[, Glp1r_neurons]
length(Cells(filtered_Glp1r_neurons)) 
saveRDS(filtered_Glp1r_neurons, "filtered_Glp1r_neurons.rds")
```

##filter neurons to only gfral and not glp1r
```{r}
threshold <- 0
# Identify cells where at least one of the genes exceeds the threshold (Gfral only)
Gfral_neurons <- rownames(gene_expression)[gene_expression$Glp1r == 0 & gene_expression$Gfral > threshold]
filtered_Gfral_neurons <- neurons[, Gfral_neurons]
length(Cells(filtered_Gfral_neurons)) 
saveRDS(filtered_Gfral_neurons, "filtered_Gfral_neurons.rds")
```